---
title: "walt_djfmp_explore"
format: html
---

## Tidying Data:

```{r}
#loading packages
library(ggplot2)
library(here)
library(tidyverse)
library(dplyr)

```

```{r}
#reading data
sample_sites <- read.csv(here("data","processed","sample_sites.csv"))
taxonomy <- read.csv(here("data","processed","taxonomy.csv"))
seine <- read.csv(here("data","processed","seine_76_25.csv"))
trawl_02_25 <- read.csv(here("data","processed","trawl_02_25.csv"))
trawl_76_01 <- read.csv(here("data","processed","trawl_76_01.csv"))
```

```{r}
#Separate sample date into day, month year (thanks abby for the code)
trawl_76_01 <- trawl_76_01 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_02_25 <- trawl_02_25 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

seine <- seine %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_76_01 <- trawl_76_01 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

trawl_02_25 <- trawl_02_25 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

seine <- seine %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )


combo_trawl <- bind_rows(trawl_76_01, trawl_02_25)
combo_trawl$year <- as.numeric(combo_trawl$year)

```

### Joining Tables for Taxonomy

```{r}
#combining the trawl data with taxonomy
trawl_joined <- combo_trawl %>% 
  left_join(taxonomy, by = "OrganismCode")
  
```

```{r}
#joing seine and toxonmy tables
seine_tax <- seine %>%
  left_join(taxonomy, by = "OrganismCode")

```

### Totaling Counts by location, year, and Species

```{r}
#total counts of each spps for each location for each year
combo_trawl_counts <- combo_trawl %>% 
  group_by(Location, year, CommonName) %>% #selecting location year and species
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop") %>% #totaling the values in the counts column for each group
  filter(CommonName != "No catch") #removing observationgs with no spp caught
```

```{r}
seine_counts <- seine_tax %>% 
  group_by(Location, year, OrganismCode) %>% #selecting location year and species
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop") %>% #totaling the values in the counts column for each group
  filter(OrganismCode != "NOFISH") #removing observationgs with no spp caught
```

### Totaling Counts of Unique Species In Seine Data

```{r}
#grouping seine data counts for native status at each location and year
seine_richness <- seine_tax %>% 
  select(Location, year, OrganismCode, CommonName.x, Family, NonNative) %>% 
  group_by(Location, year, OrganismCode, NonNative) %>% 
  summarize(count = n(), .groups = "drop") %>% #creates a column that totals the columns of that loc, year, and organism
  filter(OrganismCode != "NOFISH") %>% 
  group_by(Location, year, NonNative) %>% 
  summarize(count = n(), .groups = "drop") # allows us to total the unqiue number of organisms in each location each year seperated by native status
```

## 1 - Trawl Abundance Counts

### 1.1 Spps counts by location through time (trawl)

```{r}
#plot that shows the counts of spps for each locaton through time
#EXPLORITORY USE ONLY
count_year <- ggplot(data = combo_trawl_counts, mapping = aes(x = year, y = Count)) +
  geom_point() +
  facet_wrap(Location ~ .) +
  ylim(0,250000) #helps ignore a couple of outliers for visual scaling purposes

count_year
```

### 1.2a Counts of abundant spp through time (trawl)

```{r warning = FALSE, message = FALSE}
#plot for visualzing the most abundant spps through time
count_species <- combo_trawl_counts %>%
  group_by(CommonName) %>%
  filter(n() >= 50) %>% #only includes spps that were observed with more than 50 observations to help limit too many facets
  ungroup() %>%
  ggplot(aes(x = year, y = Count)) +
  geom_point(aes(color = Location)) +
  geom_line(aes(color = Location)) +
  facet_wrap(CommonName ~ ., scales = "free_y")

count_species
```

### 1.2b Counts of abundant families through time (trawl)

```{r warning = FALSE, message = FALSE}
#EXPLORITIVE USE ONLY
#creating a data table that organizes spps into phylogenic families
trawl_counts_taxonomy <- trawl_joined %>% 
  filter(CommonName.x != "No catch",Family != "NA") %>% 
  group_by(Location, year, Family) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups= "drop") 

#creating plot that shows abundance of families through time
count_family <- trawl_counts_taxonomy %>% 
  group_by(Family) %>%
  filter(n() >= 50) %>% #only includes families that were observed with more than 50 observations to help limit too many facets
  ungroup() %>%
  ggplot(aes(x = year, y = Count)) +
  geom_point(aes(color= Location)) +
  geom_line(aes(color= Location)) +
  facet_wrap(Family ~ ., scales = "free_y")

count_family
```

### 1.3 Counts of abundance counts through time (trawl, locations merged)

```{r warning = FALSE, message = FALSE}
#Counts of all spps through time (locations combined)
#EXPLORITORY USE ONLY
count_total <- ggplot(data = combo_trawl_counts, mapping = aes(x = year, y = Count)) +
  geom_point(aes(color=CommonName)) +
  geom_smooth(aes(color=CommonName)) +
  ylim(0,100000) + #limiits outlier scaling issues
  theme(legend.position = "none")

count_total
```

### 1.4 Native vs Invasive

```{r warning = FALSE, message = FALSE}
#EXPLORATION USE ONLY

#creating data frame that organizes into native status 
native_trawl <- trawl_joined %>% 
   filter(CommonName.x != "No catch", NonNative != "NA") %>% #removes any observations with no catch
  group_by(Location, year, Family, NonNative) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop") 

#creates a stacked area plot of abundances of native and non-native species, seperated by each location
native_area <- ggplot(data = native_trawl, mapping = aes(x = year, y = Count)) +
  geom_area(aes(fill= NonNative)) +
  facet_wrap(Location ~ .) +
  scale_y_log10()

native_area

```

## 2 - Trawl Data on Focal Species

look at how anchovies salmonids chinnok coorelate over time by location and by temperature.

### 2.1 Counts of focal salmonids per year 

```{r}
#creating a filtered data table with counts of salmonids 
focal_trawl_counts <- combo_trawl %>% 
  filter(OrganismCode == "CHN"| OrganismCode ==  "RBT") %>% #filters for rows with chinook and steelhead
  filter(StationCode == "SB018M" | 
        StationCode == "SB018N" |
        StationCode == "SB018S" | 
        StationCode == "SB018X" |
        StationCode == "SJ054M" |
        StationCode == "SR055M") %>% #filters for the specific stations that have the most long term data
  group_by(StationCode, year, OrganismCode) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")
  
#plot for abundance of chinook and steelhead, colored by station
salmon_spp_station <- ggplot(data= focal_trawl_counts, aes(x = year, y = Count, fill = StationCode)) +
  geom_line(aes(color = StationCode)) +
  facet_wrap(OrganismCode ~., scales = "free_y")

#plot of abundance of chinook and steelhead, all stations combined
salmon_spp <- ggplot(data= focal_trawl_counts, aes(x = year, y = Count )) +
  geom_line() +
  facet_wrap(OrganismCode ~., scales = "free_y")

salmon_spp_station
salmon_spp
```

### 2.2 Counts of focal salmonids per year (water temp overlay)

```{r}
#creating data frame that includes annual average water temperature 
focal_filter2 <- combo_trawl %>% 
  filter(OrganismCode == "CHN"| OrganismCode ==  "RBT") %>% #filters for rows with chinook and steelhead
  filter(StationCode == "SB018M" | 
        StationCode == "SB018N" |
        StationCode == "SB018S" | 
        StationCode == "SB018X" |
        StationCode == "SJ054M" |
        StationCode == "SR055M") %>% #filters for the specific stations that have the most long term data
  group_by(year) %>% 
  mutate(avg_temp = mean(WaterTemp, na.rm = TRUE), .groups = "drop") %>% #creating a mean on of temperature from all oberservatons that yar
  group_by(StationCode, year, OrganismCode, avg_temp) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")
```

```{r}
#plot for temperature on salmon populations through time
salmon_spp_station <- ggplot(data= focal_filter2, aes(x = year, y = Count, color = avg_temp)) +
  geom_line() +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap(OrganismCode ~., scales = "free_y",
             labeller = labeller(
               OrganismCode = c(
                 "CHN" = "Chinook Salmon",
                 "RBT" = "Steelhead"))) +
  theme_classic()+
  labs(
  title = "Salmon Population Counts Through Time",
  x = "Year",
  y = "Population Count",
  color = "Avg Temp (°C)"
)
  
salmon_spp_station
```

```{r}
#export plot as a .png into the output folder
ggsave(
  filename = "salmontemp.png",
  plot = salmon_spp_station,      # optional if it's the last plot
  width = 8,           # in inches
  height = 4,          # in inches
  dpi = 300,            # resolution (300 = publication quality)
  path = "../outputs"
)
```

### 2.3 Abundances and Temp relationship (trawl)

```{r}
#plot of relatiionship between salmon abundances and average annual temperature
salmon_spp_station2 <- ggplot(data= focal_filter2, aes(x = avg_temp, y = Count)) +
  geom_point() +
  facet_wrap(OrganismCode ~., scales = "free_y") +
  theme_classic()
salmon_spp_station2
```

### 2.4 Counts of focal salmonids per year (water flow overlay, trawl)

```{r}
#creating data frame that includes annual average water flow 
focal_filter_flow <- combo_trawl %>% 
  filter(OrganismCode == "CHN"| OrganismCode ==  "RBT") %>% #filters for rows with chinook and steelhead
  filter(StationCode == "SB018M" | 
        StationCode == "SB018N" |
        StationCode == "SB018S" | 
        StationCode == "SB018X" |
        StationCode == "SJ054M" |
        StationCode == "SR055M") %>% #filters for the specific stations that have the most long term data
  group_by(year) %>% 
  mutate(avg_flow = mean(FlowmeterDifference, na.rm = TRUE), .groups = "drop") %>% #creating a mean of flow from all oberservatons that year
  group_by(StationCode, year, OrganismCode, avg_flow) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")
```

```{r}
#plot for flow on salmon populations through time
salmon_spp_station_flow <- ggplot(data= focal_filter_flow, aes(x = year, y = Count, color = avg_flow)) +
  geom_line() +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap(OrganismCode ~., scales = "free_y")

salmon_spp_station_flow
```

### salmonids over years (thanks for the base code abbie!)

```{r}
salmonids_by_year <- combo_trawl %>%
  filter(!is.na(CommonName)) %>%
  filter(CommonName == "Chinook Salmon" | CommonName == "steelhead") %>% 
  group_by(year, CommonName) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(year, desc(count))

salmonids_by_year

#Chinook by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="Chinook Salmon"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

#Steelhead by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="steelhead"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

```

### looking at avg water flow by year (adjusted for flow from temp code written by abbie)

```{r}
#Cal avg water temp by year
avg_flow_by_year <- combo_trawl %>%
  group_by(year) %>%
  summarise(avg_flow = mean(FlowmeterDifference, na.rm = TRUE),
            .groups = "drop")

#Filter for chinook & steelhead individually
ch_by_year <- salmonids_by_year %>% 
  filter(CommonName == "Chinook Salmon")
sh_by_year <- salmonids_by_year %>% 
  filter(CommonName == "steelhead")

#flow/ salmon plot data
flow_ch_data <- left_join(ch_by_year, avg_flow_by_year, by = "year")
flow_sh_data <- left_join(sh_by_year, avg_flow_by_year, by = "year")

#Scale for dual axis plotting chinook
scale_factor_ch <- max(flow_ch_data$count, na.rm = TRUE) / 
                max(flow_ch_data$avg_flow, na.rm = TRUE)
flow_ch_data <- flow_ch_data %>%
  mutate(flow_scaled = avg_flow * scale_factor_ch)

#Scale for dual axis plotting steelhead
scale_factor_sh <- max(flow_sh_data$count, na.rm = TRUE) / 
                max(flow_sh_data$avg_flow, na.rm = TRUE)
flow_sh_data <- flow_sh_data %>%
  mutate(flow_scaled = avg_flow * scale_factor_sh)


#Dual axis plot chinook 
ggplot(flow_ch_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = flow_scaled, color = "Avg Water Flow"), 
            linetype = "solid", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor_ch,
                        name = "Average Water Flow")
  ) +
  
  theme_minimal() 

#Dual axis plot steelhead 
ggplot(flow_sh_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = flow_scaled, color = "Avg Water Flow"), 
            linetype = "solid", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor_sh,
                        name = "Average Water Flow")
  ) +
  
  theme_minimal() 


```
