---
title: "Ward_DJFMP_explore"
author: "Ward, A."
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls()) #clean environment

library(here)
knitr::opts_knit$set(
  root.dir = here::here()
)

knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r packages, include=FALSE}
library(readr)
library(tidyverse)
library(readxl) 
library(dplyr)
library(janitor) 
library (stringr)
library(openxlsx)
library(stats)
library(car)
library(ggplot2)
library(ggpubr)
library(lubridate)
```

#Import Data
```{r loading in data}
#loading in data
sample_sites <- read_csv("./data/processed/sample_sites.csv")
taxonomy <- read_csv("./data/processed/taxonomy.csv")
trawl_02_25 <- read_csv("./data/processed/trawl_02_25.csv")
trawl_76_01 <- read_csv("./data/processed/trawl_76_01.csv")
```

#Data Cleaning and Tidying
```{r}
#Separate sample date into day, month year
trawl_76_01 <- trawl_76_01 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_02_25 <- trawl_02_25 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_76_01 <- trawl_76_01 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

trawl_02_25 <- trawl_02_25 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

combo_trawl <- bind_rows(trawl_76_01, trawl_02_25)
combo_trawl$year <- as.numeric(combo_trawl$year)

```


#Long term Trends of Species

First, I am going to explore long term trends of the data, abundance of species, environmental trends etc. This will help to figure out a direction to go forward with.  

##Total Abundance Shifts over time
Ok first want to look at total abundance has changed over time for all species 
```{r}
annual_abundance <- combo_trawl %>%
  group_by(year) %>%
  summarise(total_fish = sum(Count, na.rm = TRUE))

ggplot(annual_abundance, aes(year, total_fish)) +
  geom_line() +
  theme_minimal()
```
Interestingly a huge spike at the beginning of sampling (sampling effort?) with highs and lows throughout the rest of the years sampled. What would happen if we filter out that first peak?

```{r}
ggplot(subset(annual_abundance, year >1978 ), aes(year, total_fish)) +
  geom_line() +
  theme_minimal()
```
Interesting to see certain peaks more clearly.

##Species richenss changes over time
```{r}
annual_richness <- combo_trawl %>%
  group_by(year) %>%
  summarise(richness = n_distinct(CommonName))

ggplot(annual_richness, aes(year, richness)) +
  geom_line() +
  theme_minimal()
```

##Most abundant species

What are the top 10 most abundant species?
```{r}
top_species <- combo_trawl %>%
  group_by(CommonName) %>%
  summarise(total = sum(Count, na.rm = TRUE)) %>%
  arrange(desc(total)) %>%
  slice(1:10) %>%
  pull(CommonName)

print(top_species)
```


##Unique species over the years
I want to first see how many unique species there are in all of the trawl data. Then I want to look at the abundace of each individual species by year 
```{r}
unique(combo_trawl$CommonName)

#Species
species_by_year <- combo_trawl %>%
  filter(!is.na(CommonName)) %>%
  group_by(year, CommonName) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(year, desc(count))

species_by_year

ggplot(species_by_year, aes(x = year, y = count, color = CommonName)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  theme(
    legend.position = "none")


```
Abundances are so variable by species this is difficult to tell honestly. The magnitude is so different. 


#Long Term Trends -- Environmental Variables 

##Looking at seasonal water temp differences
```{r}
ggplot(combo_trawl, aes(x = month, y = WaterTemp)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  labs(
    x = "Month",
    y = "Water Temperature"
  ) +
  theme_minimal()

seasonal_means <- combo_trawl %>%
  group_by(month) %>%
  summarise(
    avg_WaterTemp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(seasonal_means, aes(x = month, y = avg_WaterTemp, group = 1)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    title = "Average Seasonal Cycle of Water Temperature",
    x = "Month",
    y = "Average Water Temperature"
  ) +
  theme_minimal()


combo_trawl <- combo_trawl %>%
  mutate(
    season = case_when(
      month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      month %in% c("Mar", "Apr", "May") ~ "Spring",
      month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      month %in% c("Sep", "Oct", "Nov") ~ "Fall"
    )
  )

ggplot(combo_trawl, aes(x = season, y = WaterTemp)) +
  #facet_wrap(~year) +
  geom_boxplot(fill = "tan", alpha = 0.7)

```

##Are seasonal water temps sig different?
```{r}
seasonal_water_temp_avg <- combo_trawl %>%
  group_by(year, season) %>%
  summarise(
    avg_WaterTemp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

anova_waterseason1 <- aov(avg_WaterTemp ~ season, data = seasonal_water_temp_avg)
summary(anova_waterseason1)
```

##How has water temp changed over time 

```{r}
annual_water_temp <- combo_trawl %>%
  group_by(year) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE)
  )

ggplot(annual_water_temp, aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  theme_minimal()
```
Let's look at how trends have changes over years
```{r}
#Changing the sample date column to recognize as date - ugh
combo_trawl$SampleDate <- as.Date(combo_trawl$SampleDate, format = "%m/%d/%Y")

#Get DOY for mapping
combo_trawl <- combo_trawl %>%
   mutate(DOY = yday(SampleDate))

#avg temp per day since multiple sampling per day
daily_water_temp <- combo_trawl %>%
  group_by(year, DOY) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(daily_water_temp, aes(x = DOY, y = mean_temp, group = year, color = factor(year))) +
  geom_line(alpha = 0.7) +
  theme_minimal()

ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_line(color = "gray70", alpha = 0.5) +
  geom_line(
    data = subset(daily_water_temp, year >= 2020),
    aes(color = factor(year)),
    size = 1
  ) +
  theme_minimal()

```
Ok interesting but a little crazy, let's try by decade. I want to see if peaks or falls are happening sooner or later than previously.

```{r}

#Get decade for mapping
combo_trawl <- combo_trawl %>%
  mutate(
    decade = floor(year / 10) * 10
  )

decade_water_temp <- combo_trawl %>%
  group_by(decade, DOY) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(decade_water_temp, aes(DOY, mean_temp, color = factor(decade))) +
  geom_line(size = 0.7) +
  theme_minimal()

anova_decadaltemps1 <- aov(mean_temp ~ decade, data = decade_water_temp)
summary(anova_decadaltemps1)
```


##Ok 70s and 80s looks like they didn't really sample in the summer months? Is that true?
```{r}
# Count number of samples per decade and month
sampling_coverage <- combo_trawl %>%
  group_by(decade, month) %>%
  summarise(samples = n(), .groups = "drop")

# Plot sampling frequency
ggplot(sampling_coverage, aes(x = month, y = samples, fill = factor(decade))) +
  geom_col(position = "dodge") 
```
Ok so 1990 and on have the most sampling --  maybe we should focus on those?

##Flow Patterns Seasonally
```{r}
#Daily flow patterns (I think)
daily_flow <- combo_trawl %>%
  group_by(year, season, SampleDate) %>%
  summarise(
    daily_flow = mean(FlowmeterDifference, na.rm = TRUE),
    .groups = "drop"
  )

#Seasonal flow averages
seasonal_flow <- daily_flow %>%
  group_by(year, season) %>%
  summarise(
    mean_flow = mean(daily_flow, na.rm = TRUE),
    max_flow  = max(daily_flow, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(seasonal_flow, aes(x = season, y = mean_flow, fill = season)) +
  geom_boxplot() +
  theme_minimal() 
```
##FLow Patterns Annually
```{r}
annual_flow <- seasonal_flow %>%
  group_by(year) %>%
  summarise(total_flow = sum(mean_flow, na.rm = TRUE))

ggplot(annual_flow, aes(x = year, y = total_flow)) +
  geom_line(color = "blue") +
  geom_point() +
  theme_minimal()
```


#Salmonid Focus
Now I want to focus specifically on salmonids

##salmonids over years
```{r}
salmonids_by_year <- combo_trawl %>%
  filter(!is.na(CommonName)) %>%
  filter(CommonName == "steelhead" | CommonName == "Chinook Salmon") %>% 
  group_by(year, CommonName) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(year, desc(count))

salmonids_by_year

#Chinook by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="Chinook Salmon"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

#Steelhead by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="steelhead"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

```
##looking at avg water temp by year
```{r}
#Cal avg water temp by year
avg_temp_by_year <- combo_trawl %>%
  group_by(year) %>%
  summarise(avg_temp = mean(WaterTemp, na.rm = TRUE),
            .groups = "drop")

#Filter for chinook & steelhead individually
ch_by_year <- salmonids_by_year %>% 
  filter(CommonName == "Chinook Salmon")
sh_by_year <- salmonids_by_year %>% 
  filter(CommonName == "steelhead")

#temp/ salmon plot data
temp_ch_data <- left_join(ch_by_year, avg_temp_by_year, by = "year")
temp_sh_data <- left_join(sh_by_year, avg_temp_by_year, by = "year")

#Scale for dual axis plotting chinook
scale_factor_ch <- max(temp_ch_data$count, na.rm = TRUE) / 
                max(temp_ch_data$avg_temp, na.rm = TRUE)
temp_ch_data <- temp_ch_data %>%
  mutate(temp_scaled = avg_temp * scale_factor_ch)

#Scale for dual axis plotting steelhead
scale_factor_sh <- max(temp_sh_data$count, na.rm = TRUE) / 
                max(temp_sh_data$avg_temp, na.rm = TRUE)
temp_sh_data <- temp_sh_data %>%
  mutate(temp_scaled = avg_temp * scale_factor_sh)


#Dual axis plot chinook 
ggplot(temp_ch_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = temp_scaled, color = "Avg Water Temp"), 
            linetype = "solid", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor_ch,
                        name = "Average Water Temperature")
  ) +
  
  theme_minimal() 

#Dual axis plot steelhead 
ggplot(temp_sh_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = temp_scaled, color = "Avg Water Temp"), 
            linetype = "solid", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor_sh,
                        name = "Average Water Temperature")
  ) +
  
  theme_minimal() 


```





#Looking at avg water temp by year and location
```{r}
avg_temp_table <- combo_trawl %>%
  group_by(year, Location) %>%
  summarise(
    avg_WaterTemp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

avg_temp_table


ggplot(avg_temp_table, aes(x = as.numeric(year), y = avg_WaterTemp, color = Location)) +
  geom_point(size = 2) +
  geom_line() +
   geom_smooth(
      aes(group = 1),
      method = "lm",
      se = TRUE,
      color = "black",
      linewidth = 1,
      linetype = "dashed"
    ) +
  labs(
    title = "Average Water Temperature Over Time by Location",
    x = "Year",
    y = "Average Water Temperature"
  ) +
  theme_minimal()


```



#Let's start map exploration 
Basic Map of Sample Sites
```{r}
library(sf)
library(terra)
library(dplyr)
library(spData)
#install.packages("spDataLarge", repos = "https://nowosad.github.io/drat/", type = "source")

library(spDataLarge)
#install.packages("tmap", repos = c("https://r-tmap.r-universe.dev","https://cloud.r-project.org"))
library(tmap)    # for static and interactive maps
library(leaflet) # for interactive maps
library(ggplot2) # tidyverse data visualization package
library(ggthemes)
library(rnaturalearth) #more base maps

library(osmdata) #waterway data
library(elevatr) #elevations
library(nhdplusTools) #NHD datasets

```
##BB of coordinates of delta
```{r}
# Bounding box coordinates (min lon, min lat, max lon, max lat)
delta_bbox <- st_bbox(c(xmin = -121.25, ymin = 37.47, xmax = -122.41, ymax = 39.37),
                      crs = st_crs(4326))

delta_bbox_sf <- st_as_sfc(delta_bbox)
```

##Mapping water
```{r}
#waterways
water_osm <- opq(bbox = delta_bbox) %>%
  add_osm_feature(key = "waterway") %>%
  osmdata_sf()

# extract rivers, canals, etc
rivers <- water_osm$osm_lines
```

```{r}
#lakes 
lakes_osm <- opq(bbox = delta_bbox) %>%
  add_osm_feature(key = "natural", value = "water") %>%
  add_osm_feature(key = "water",
                  value = c("lake", "reservoir", "bay", "lagoon"),
                  match_case = FALSE) %>%
  osmdata_sf()

water_polygons <- lakes_osm$osm_polygons
```

##Map of waterways
```{r}
ggplot() +
  geom_sf(data = st_crop(water_polygons, delta_bbox),
          fill = "lightblue", color = NA) +
  geom_sf(data = st_crop(rivers, delta_bbox),
          color = "blue", size = 0.3) +
  coord_sf(xlim = c(delta_bbox$xmin, delta_bbox$xmax),
           ylim = c(delta_bbox$ymin, delta_bbox$ymax)) +
  theme_minimal()
```

```{r}
sample_sites_sf <- st_as_sf(
  sample_sites,
  coords = c("Longitude", "Latitude"),
  crs = 4326
) 
```

##Map of sample sites
```{r}
ggplot() +
  geom_sf(data = st_crop(water_polygons, delta_bbox),
          fill = "lightblue", color = NA) +
  geom_sf(data = st_crop(rivers, delta_bbox),
          color = "blue", size = 0.3) +
  coord_sf(xlim = c(delta_bbox$xmin, delta_bbox$xmax),
           ylim = c(delta_bbox$ymin, delta_bbox$ymax)) +
  geom_sf(data = sample_sites_sf, color = "red", size = 2) +
  theme_minimal()
```
Ok that was difficult! I would like to find a way to reduce the amount of water shown, focusing primarily on bays estuaries and major rivers and figure out a better bb for the delta.

Moving on...


#Separating Sampling Methods
Ok only going to focus on salmonids
```{r}
#Unique
unique(combo_trawl$MethodCode)

#Filtering for salmonids and for specific sites  
salmonid_trawl <- combo_trawl %>% 
  filter(OrganismCode == "CHN" | OrganismCode == "RBT") 
  #filter(StationCode == "SB018M" | StationCode == "SB018N" | StationCode == "SB018S" |StationCode == "SB018X" | StationCode == "SJ054M" |StationCode == "SR055M")

#Summming the counts per year to get abundances and getting runs 
salmonid_trawl <- salmonid_trawl %>% 
  mutate(
    run = case_when(
      OrganismCode == "RBT" ~ "steelhead",
      !is.na(RaceByLength) ~ RaceByLength,
      !is.na(RaceByTag) ~ RaceByTag,
      TRUE ~ NA_character_
    )
  )

salmonid_trawl <- salmonid_trawl %>%
  mutate(date = as.Date(SampleDate))
```

## Annual Abundance
```{r}
#Annual abundaces all salmonids
ggplot(
  salmonid_trawl %>%
    group_by(year) %>%
    summarise(annual_abundance = sum(Count, na.rm = TRUE),
              .groups = "drop"),
  aes(x = year,
      y = annual_abundance)
) +
  geom_line(color = "black") +
  geom_smooth(method = "lm")+
  geom_point() +
  labs(
    x = "Year",
    y = "Total Annual Salmonid Abundance",
    title = "Annual Salmonid Abundance Over Time"
  ) +
  theme_bw()

ggsave(
  filename = "/Users/abigailward/Downloads/annual_salmonid_abundance.png",
  width = 8,
  height = 6,
  dpi = 300
)

#Add annual abundance of fish column
salmonid_trawl <- salmonid_trawl %>%
  group_by(year) %>%
  mutate(annual_abundance = sum(Count, na.rm = TRUE)) %>%
  ungroup()

#Create annual abundace dataset
annual_data <- salmonid_trawl %>%
  group_by(year) %>%
  summarise(
    annual_abundance = sum(Count, na.rm = TRUE),
    .groups = "drop"
  )

#See how salmon aubdance is changing over time 
glm(annual_abundance ~ year,
    family = poisson,
    data = annual_data)



#Annual abundaces by species 
ggplot(
  salmonid_trawl %>%
    group_by(year, OrganismCode) %>%
    summarise(annual_abundance = sum(Count, na.rm = TRUE),
              .groups = "drop"),
  aes(x = year,
      y = annual_abundance,
      color = OrganismCode)
) +
  geom_line() +
  geom_point() +
  theme_bw()

#annual abundaces by run
ggplot(
  salmonid_trawl %>%
    group_by(year, OrganismCode, run) %>%
    summarise(annual_abundance = sum(Count, na.rm = TRUE),
              .groups = "drop"),
  aes(x = year,
      y = annual_abundance,
      color = run)
) +
  geom_line() +
  facet_wrap(~ OrganismCode) +
  theme_bw()

```
```{r}
exp(-0.01146)
```


About a 1.1% decline per year on average since salmon abundance is mult by 0.9886. 

Negative Binomial Generalized Linear Model??

```{r}
library(MASS)

model_nb <- glm.nb(annual_abundance ~ year,
                   data = annual_data)

summary(model_nb)
```

```{r}
exp(-0.01571)
```

About a 1.56% decline per year.


## Water temps only for salmonid stations 
```{r}
ggplot(salmonid_trawl, aes(x = month, y = WaterTemp)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  labs(
    x = "Month",
    y = "Water Temperature"
  ) +
  theme_minimal()

ggsave(
  filename = "/Users/abigailward/Downloads/seasonal_watertemp.png",
  width = 8,
  height = 6,
  dpi = 300
)

ggplot(salmonid_trawl, aes(x = season, y = WaterTemp)) +
  #facet_wrap(~year) +
  geom_boxplot(fill = "tan", alpha = 0.7)


#How has water temp changed over time
annual_water_tempsalmo <- salmonid_trawl %>%
  group_by(year) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE)
  )

ggplot(annual_water_temp, aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  labs(
    x = "Year",
    y = "Mean Water Temperature"
  ) +
  theme_bw()

ggsave(
  filename = "/Users/abigailward/Downloads/annual_watertemp1.png",
  width = 8,
  height = 6,
  dpi = 300
)

annual_water_temp %>% 
  filter(year > 1995) %>% 
  ggplot(aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  labs(
    x = "Year",
    y = "Mean Water Temperature"
  ) +
  theme_bw()

ggsave(
  filename = "/Users/abigailward/Downloads/annual_watertemp2.png",
  width = 8,
  height = 6,
  dpi = 300
)
```

##CPUE
```{r}
salmonid_trawl <- salmonid_trawl %>%
  mutate(CPUE = Count / Volume) 

cpue_summary <- salmonid_trawl %>%
  group_by(year, CommonName, run) %>%
  summarise(
    mean_CPUE = mean(CPUE, na.rm = TRUE),
    total_fish = sum(Count, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(cpue_summary, aes(x = year, y = mean_CPUE, color = CommonName)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_grid(~CommonName, scales = "free_y") +
  theme_bw() 

ggplot(cpue_summary, aes(x = year, y = mean_CPUE, color = run)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  theme_bw()

```

##Flow compared to abundance
```{r}
#Load in flow data
library(dplyr)
flow_70_83 <- read_csv("./data/processed/dayflow-results-1970-1983.csv")
flow_84_96 <- read_csv("./data/processed/dayflow-results-1984-1996.csv")
flow_97_23 <- read_csv("./data/processed/dayflow-results-1997-2023.csv")
water_year <- read_xlsx("./data/processed/cdec-water-year-type-jun-2025.xlsx")

water_year <- water_year %>%
  rename(Year = WY) %>% 
  filter(Area == "Sacramento Valley")

allflow <- bind_rows(flow_70_83, flow_84_96, flow_97_23)
allflow <- allflow %>% 
 rename(SampleDate = Date) 

#make sure date is date and in correct format
allflow <- allflow %>%
  mutate(SampleDate = as.Date(SampleDate, format = "%m/%d/%Y"))

#Get water year for all NAs
allflow <- allflow %>%
  left_join(
    water_year %>% dplyr::select(Year, WYT),
    by = "Year"
  ) %>%
  mutate(
    Y = WYT  
  ) %>%
  dplyr::select(-WYT) 

 
```

These datasets give net Delta outflow at Chipps island as well as inflow rates for various pathways into the Delta. For the purposes of this project, I'm focusing on flow rate out of the Delta (OUT).

## Flow and WY for selected years of DJFMP
```{r}
#annual flow for Delta
annual_flow <- allflow %>%
  group_by(Year, Y) %>%                               
  summarise(
    mean_OUT = mean(OUT, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(annual_flow, aes(x = Year, y = mean_OUT, fill = Y)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  scale_x_continuous(
    breaks = seq(min(annual_flow$Year),
                 max(annual_flow$Year),
                 by = 5)
  ) +
  labs(
    x = "Year",
    y = "Mean Annual Flow (cfs)"
  ) +
  theme_bw()

ggsave(
  filename = "/Users/abigailward/Downloads/annual_flow_WYT.png",
  width = 8,
  height = 6,
  dpi = 300
)




ggplot(allflow, aes(x = SampleDate, y = OUT)) +
  geom_line(alpha = 0.3) +           
  geom_smooth(method = "loess") +    
  theme_bw()

ggplot(annual_flow, aes(x = Year, y = mean_OUT)) +
  geom_line(alpha = 0.3) +           
  geom_smooth(method = "lm") +    
  theme_bw()
```


Water year type (Y) definitions by the state water board: 
- W : Wet year type
- AN : Above normal year type
- BN : Below normal year type
- D : Dry year type
- C : Critical year type 

Now I want to look at salmon abundaces as a function of flow:
```{r}
#Caluclate daily abundace 
daily_salmonid_abundance <- salmonid_trawl %>%
  group_by(SampleDate, CommonName, run) %>%
  summarise(
    total_fish = sum(Count, na.rm = TRUE),
    .groups = "drop"
  ) 



#Merge abundances back in 
salmonid_trawl <- salmonid_trawl %>%
  left_join(daily_salmonid_abundance,
            by = c("SampleDate", "CommonName", "run"))


#Merge with flow data
salmonid_trawl <- salmonid_trawl %>%
  left_join(
    allflow %>% dplyr::select(SampleDate, OUT, Y, Year),
    by = "SampleDate"
  )


ggplot(salmonid_trawl, aes(x = OUT, y = total_fish, color = run)) +
  geom_point(alpha = 0.5) +  
  theme_bw()

#Get annual abundance data merged with flow
annual_data<-annual_data %>% 
  rename(Year = year)

#Join dataframes
annual_data<-annual_data %>% 
  left_join(
    annual_flow, by = "Year"
  )

#GEt scale for adding salmon abundance to flow chart
scale_factor_salmo <- max(annual_data$annual_abundance, na.rm = TRUE) / 
                max(annual_data$mean_OUT, na.rm = TRUE)
annual_data <- annual_data %>%
  mutate(count_scaled = annual_abundance * scale_factor_salmo)


#Plot with both abundace and flow
ggplot(annual_data, aes(x = Year, y = mean_OUT, fill = Y)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  scale_x_continuous(
    breaks = seq(min(annual_flow$Year),
                 max(annual_flow$Year),
                 by = 5)
  ) +
  geom_line(aes(y = count_scaled, color = "Annual Salmonid Abundance", group = 1), 
            linetype = "solid", color ="black", size = 1) +
  scale_y_continuous(
    name = "Salmonid Abundance",
    sec.axis = sec_axis(~ . / scale_factor_salmo,
                        name = "Annual Salmonid Abundance")
  ) +
  labs(
    x = "Year",
    y = "Mean Annual Flow (cfs)"
  ) +
  theme_bw()

ggsave(
  filename = "/Users/abigailward/Downloads/annual_flow_abundance.png",
  width = 8,
  height = 6,
  dpi = 300
)

```

Plots of flow and temperature as a function of salmon abundance
```{r}
#Flow
ggplot(salmonid_trawl, aes(x = OUT, y = Count, color = run)) +
  geom_point(alpha = 0.5) +  
  theme_bw()

#Temp
ggplot(salmonid_trawl, aes(x = WaterTemp, y = Count, color = run)) +
  geom_point(alpha = 0.5) +  
  theme_bw()

#Turbity
ggplot(salmonid_trawl, aes(x = Turbidity, y = Count, color = run)) +
  geom_point(alpha = 0.5) +  
  theme_bw()
```




Ok let's start modeling juvenile abundances as a function of environmental variables through this time period we have on hand. 

Response variable is going to be juvenile salmonid abundance in the form of count data
Predictor variables will be flow (using OUT from the DayFlow dataset), water temp, turbidity?, water year type... any others?

Neg. binomial because of overdispertion for turbity and flow? Ask Malin or Clavin in class tomorrow 

And then look at effects over time.


```{r}
library(brms)
m.flow <- brm(
  Count ~ log(OUT + 1) + (1 | SampleDate),
  family = negbinomial(),
  data = salmonid_trawl,
  chains = 4, cores = 4, iter = 4000,
  algorithm = 'fullrank'
)
```

```{r}
summary(m.flow)
```


Ok this is taking forever to run

another approach?


