---
title: "Long term water quality effects on juvenile salmon abundance in the Sacramento San-Joaquin River Delta"
author: "Abigail Ward, Ambrosio Rivera, Walt Vance, Peter Nelson"
format:
 pdf: default
editor: visual
execute:
  echo: false
  warning: false
  message: false
  results: "hide"
  fig-align: center
  fig-width: 8 
  fig-height: 6
---

## State of Juvenile Salmon in the Delta Report

Water is an important and valuable resource throughout the state of California. The Sacramento San-Joaquin River Delta is one of the most utilized water resources, providing water to nearly 27 million people across the state as well as supporting a \$1 billion/ year agricultural industry. Yet this area also contains a diversity of over 700 plant and animal species, many of which are important in recreational and community use. Understanding how to manage this resource in a way that both supports human endeavors and the vital ecosystem within this area is extremely important as we continue to face more extreme weather events.

### Salmonid Importance

Pacific salmon and steelhead are among the most ecologically, economically, and culturally significant species connected to the Sacramento–San Joaquin River Delta. The California salmon industry is valued at \$1.4 billion, contributing to nearly 23,000 jobs annually. This economy depends on healthy runs of Chinook, coho, and steelhead utilizing these waterways. The Delta plays a vital role in the life cycle of these salmonids. Around two thirds of species of California's adult Pacific salmon pass through the Delta on their way upstream to spawn, while juveniles utilize the area as rearing habitat. This transitional zone provides an abundance of food, refuge and ideal conditions for physiological adaptations as they prepare for the marine environment.

However, salmonid populations face pressure from changes to flow regimes, water temperatures, habitat loss, and extreme weather events. Reduced snow melt runoff, prolonged drought, and flood events have disrupted migration timing and reproductive success. Effective Delta management must then balance water supply with environmental needs, habitat restoration, and temperature control measures to sustain both salmon populations and the communities that depend on them.

### Report Objectives

This assessment aims to evaluate the current state of juvenile salmonid abundances in the Sacramento–San Joaquin River Delta and how these abundances have changed over time as an effect of ecological parameters linked to water management practices. This report focuses specifically on steelhead and Chinook salmon which rely on the Delta as critical migratory and rearing habitat.

To conduct our analyses, we utilize long-term monitoring data from the Delta Juvenile Fish Monitoring Program (DJFMP). This dataset provides multi-decadal information on juvenile salmonid abundance, allowing for evaluation of temporal trends and potential relationships with flow, temperature, and other ecological parameters. Through this, the study seeks to better understand how water management strategies may influence salmonid population dynamics within the Delta and propose how water resource managers may better work in the Delta.

### Results

#### Salmonid Abundance

```{r}
#Load packages
library(here)
library(readr)
library(tidyverse)
library(readxl) 
library(dplyr)
library(janitor) 
library (stringr)
library(openxlsx)
library(stats)
library(car)
library(ggplot2)
library(ggpubr)
library(lubridate)
library(MASS)
library(brms)
library(chron)
library(gridExtra)
library(dataRetrieval)
remotes::install_github("doi-usgs/EflowStats@v5.2.0")
library(EflowStats)
library(fishualize)
library(tidybayes)
```

```{r}
#Load in data 
sample_sites <- read_csv(here("data", "processed", "sample_sites.csv"))
taxonomy <- read_csv(here("data", "processed", "taxonomy.csv"))
trawl_02_25 <- read_csv(here("data", "processed", "trawl_02_25.csv"))
trawl_76_01 <- read_csv(here("data", "processed", "trawl_76_01.csv"))

#Separate sample date into day, month year
trawl_76_01 <- trawl_76_01 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_02_25 <- trawl_02_25 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)


trawl_76_01 <- trawl_76_01 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

trawl_02_25 <- trawl_02_25 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

#Create combination of datasets
combo_trawl <- bind_rows(trawl_76_01, trawl_02_25)
combo_trawl$year <- as.numeric(combo_trawl$year)

#Create season column
combo_trawl <- combo_trawl %>%
  mutate(
    season = case_when(
      month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      month %in% c("Mar", "Apr", "May") ~ "Spring",
      month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      month %in% c("Sep", "Oct", "Nov") ~ "Fall"
    )
  )

#Salmonid only dataset
#Filtering for salmonids and for specific sites  
salmonid_trawl <- combo_trawl %>% 
  filter(OrganismCode == "CHN" | OrganismCode == "RBT") 

#Summming the counts per year to get abundances and getting runs 
salmonid_trawl <- salmonid_trawl %>% 
  mutate(
    run = case_when(
      OrganismCode == "RBT" ~ "steelhead",
      !is.na(RaceByLength) ~ RaceByLength,
      !is.na(RaceByTag) ~ RaceByTag,
      TRUE ~ NA_character_
    )
  )


```

```{r}
#| fig-cap: "Juvenile salmon abundance over time"

#Figure 1 
fig1 <- ggplot(
  salmonid_trawl %>%
    group_by(year) %>%
    summarise(annual_abundance = sum(Count, na.rm = TRUE),
              .groups = "drop"),
  aes(x = year,
      y = annual_abundance)
) +
  geom_line(color = "black") +
  geom_smooth(method = "lm")+
  geom_point() +
  labs(
    x = "Year",
    y = "Total Annual Salmonid Abundance",
    title = "Annual Salmonid Abundance Over Time"
  ) +
  theme_bw()

fig1

#savefigure 
ggsave("./figures/fig1.png", plot = fig1, width = 8, height = 6, dpi = 300)

```

```{r}
#| include: false

#Add annual abundance of fish column
salmonid_trawl <- salmonid_trawl %>%
  group_by(year) %>%
  mutate(annual_abundance = sum(Count, na.rm = TRUE)) %>%
  ungroup()

#Create annual abundace dataset
annual_data <- salmonid_trawl %>%
  group_by(year) %>%
  summarise(
    annual_abundance = sum(Count, na.rm = TRUE),
    .groups = "drop"
  )

#Perform a negative binomial glm to look at annual abundace over the years
m.year <- glm.nb(annual_abundance ~ year,
                   data = annual_data)

saveRDS(m.year, file = "./outputs/m.year.rds")
```

```{r}
summary(model_nb)
```

```{r}
#| include: false
exp(-0.01571)
```

```{r}
#| include: false
100*(1-0.9844)
```

Overall, we've seen a 1.56% decline in juvenile salmon abundances since 1976 despite millions of hatchery releases each year. Although there have been specific peaks throughout time, this general trend in the Delta is extremely concerning as a downward trend in juvenile abundance can give insight into how changes to climate and habitat have had long term negative effects. Because juvenile survival strongly influences adult return rates, sustained declines at this life stage can translate into reduced numbers of spawning adults in subsequent years. This trend is vital to understand as we begin to explore how water management strategies can uplift and boost these populations.

```{r}
#| fig-cap: "Relative abundance based on CPUE"
#FIGURE 2
trawls <- combo_trawl %>% 
  mutate(across(c(Location:StationCode), as.factor),
         date = SampleDate,
         .keep = "unused") %>% 
  clean_names() %>% 
  group_by(date, sample_time, station_code) %>% 
  mutate(sample_id = cur_group_id()) %>% 
  ungroup() %>% 
  relocate(date, .after = station_code)

temp <-
  trawls %>% 
  group_by(sample_id, organism_code) %>% 
  summarise(total_count = sum(count, 
                              na.rm = TRUE),
            .groups = "drop") %>% 
  pivot_wider(names_from = organism_code,
              values_from = total_count,
              values_fill = 0) %>% 
  dplyr::select(sample_id, CHN, RBT) %>% 
  ungroup()
  
# totals ----
# totals for each unique trawl...
trawl_sum <-
  inner_join(trawls, temp, by = "sample_id") %>% 
  dplyr::select(sample_id, CHN, RBT,
                location, region_code, station_code, date, year, sample_time,
                method_code, water_temp) %>% 
  arrange(sample_id) %>% 
  distinct()

# effort ----
# control for effort (# of trawls)

trawl_indices <- 
  trawl_sum %>% 
  group_by(year) %>% 
  summarise(total_chn = sum(CHN),
            total_rbt = sum(RBT),
            n = n(),
            # index of abundance (total fish/number of trawls)
            Chinook = sum(CHN)/n,
            Steelhead = sum(RBT)/n) %>% 
  pivot_longer(cols = c(Chinook, Steelhead),
               names_to = "species",
               values_to = "index")

# plot ----
fig2<- ggplot(trawl_indices,
       aes(x = year, y = index, color = species)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "loess") +
  facet_wrap(~ species, ncol = 1, scales = "free_y") +
  labs(x = "Year", y = "Abundance Index", color = "Species") +
  scale_color_manual(values = c(
    "Chinook" = "steelblue",
    "Steelhead" = "darkorange"
  )) +
  theme_minimal()

fig2

#savefigure 
ggsave("./figures/fig2.png", plot = fig2, width = 8, height = 6, dpi = 300)

```

When factoring catch per unit effort based on the number of trawls per sampling site, we see an ever sharper decline in abundance over time (Fig. 2). Relative abundance of steelhead has remained significantly smaller than Chinook salmon, however, comparatively, Chinook have been trending sharply downward over time despite the amount of trawling effort. Yet, for all salmonid species, we see fewer fish detected over per unit sampling effort over time.

#### Flow Rate Trends and Effects

```{r}
#Load in flow data
flow_70_83 <- read_csv(here("data", "processed", "dayflow-results-1970-1983.csv"))
flow_84_96 <- read_csv(here("data", "processed", "dayflow-results-1984-1996.csv"))
flow_97_23 <- read_csv(here("data", "processed", "dayflow-results-1997-2023.csv"))
water_year <- read_xlsx(here("data", "processed", "cdec-water-year-type-jun-2025.xlsx"))

#Filter Water Year Data
water_year <- water_year %>%
  rename(Year = WY) %>% 
  filter(Area == "Sacramento Valley")

#Combine Flow Data
allflow <- bind_rows(flow_70_83, flow_84_96, flow_97_23)
allflow <- allflow %>% 
 rename(SampleDate = Date) 

#Correct Flow Dates
allflow <- allflow %>%
  mutate(SampleDate = as.Date(SampleDate, format = "%m/%d/%Y"))


#Get water year for all NAs
allflow <- allflow %>%
  left_join(
    water_year %>% dplyr::select(Year, WYT),
    by = "Year"
  ) %>%
  mutate(
    Y = WYT  
  ) %>%
  dplyr::select(-WYT) 

#Caluclate daily abundace 
daily_salmonid_abundance <- salmonid_trawl %>%
  group_by(SampleDate, CommonName, run) %>%
  summarise(
    total_fish = sum(Count, na.rm = TRUE),
    .groups = "drop"
  ) 


#Merge abundances back in 
salmonid_trawl <- salmonid_trawl %>%
  left_join(daily_salmonid_abundance,
            by = c("SampleDate", "CommonName", "run"))

#Correct Flow Dates
salmonid_trawl <- salmonid_trawl %>%
  mutate(SampleDate = as.Date(SampleDate, format = "%m/%d/%Y"))


#Merge with flow data
salmonid_trawl <- salmonid_trawl %>%
  left_join(
    allflow %>% dplyr::select(SampleDate, OUT, Y, Year),
    by = "SampleDate"
  )

 
```

```{r}
#| fig-cap: "Mean annual flow by water year type"

#FIGURE 3
#annual flow for Delta
annual_flow <- allflow %>%
  group_by(Year, Y) %>%                               
  summarise(
    mean_OUT = mean(OUT, na.rm = TRUE),
    .groups = "drop"
  )

#Make annual flow figure
fig3 <-ggplot(annual_flow, aes(x = Year, y = mean_OUT, fill = Y)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  scale_x_continuous(
    breaks = seq(min(annual_flow$Year),
                 max(annual_flow$Year),
                 by = 5)
  ) +
  labs(
    x = "Year",
    y = "Mean Annual Flow (cfs)"
  ) +
  theme_bw()

fig3

#savefigure 
ggsave("./figures/fig3.png", plot = fig3, width = 8, height = 6, dpi = 300)

```

```{r}
#| fig-cap: "Linear trend of annual flow"

#FIGURE 4
fig4 <- ggplot(annual_flow, aes(x = Year, y = mean_OUT)) +
  geom_line(alpha = 0.3) +           
  geom_smooth(method = "lm") + 
  labs(
    x = "Year",
    y = "Mean Annual Flow (cfs)"
  ) +
  theme_bw()

fig4

#savefigure 
ggsave("./figures/fig4.png", plot = fig4, width = 8, height = 6, dpi = 300)
```

Over the past several decades, we've seen increasingly drier conditions and a reduction in overall water availability. These conditions have severe consequences for juvenile salmonids who rely on freshwater flows through the Delta. This long-term monitoring has suggested that average flows through the Delta have trended downward over time (Fig. 4). We also have see drier conditions in the Sacramento River System as a whole, with an increase in "Dry" and "Critical" years since 2000 (Fig. 3).

However, even in years classified as "Above Normal", the amount of water flowing out of the Delta can remain relatively low due to upstream storage, diversions, and exports to Southern California, agriculture and the Bay area (Fig. 2). This means that the ecological benefits of wetter years are not always fully realized within the Delta ecosystem or by the species within. Water management decisions during these times, then, become critical to sustaining populations and shaping the ecological conditions of the Delta.

```{r}
#| fig-cap: "Daily salmonid abundance as a function of flow"

#Log flow to scale
salmonid_trawl$log_flow <- log1p(salmonid_trawl$OUT) #log1p to handle zeros

#FIGURE 5
fig5 <- ggplot(salmonid_trawl, aes(x = log_flow, y = Count, color = Y)) +
  geom_point()  +
  scale_color_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  labs(
    x = expression(log[10](Flow)),
    y = "Daily Abundance Count"
  ) +
  theme_bw()

fig5

#savefigure 
ggsave("./figures/fig5.png", plot = fig5, width = 8, height = 6, dpi = 300)

```

```{r}
#| include: false

#Model looking at the effects of flow on daily salmon abundance using a generalized additive model since non-linear effects 
library(mgcv)

m.flow <- gam(Count ~ s(log_flow),
              data = salmonid_trawl,
              family = nb())

saveRDS(m.flow, file = "./outputs/m.flow.rds")
```

```{r}
summary(m.flow)
```

Flow appears to influence salmonid abundance, with juvenile salmon showing higher counts within a certain range of flows (Fig. 5). This is well-known within salmonid ecology, that flows directly influence juvenile salmon health and abundance. To better understand this relationship, we fit a generalized additive model (GAM) to account for non linearity within the relationship, using a negative binomial distribution to account for overdispertion within the data. The smooth term (indicated by edf) for log(flow) was highly significant which indicates a strong non-linear relationship between flow and abundance, but this model only explained 2.93% of the variability within our data. This points towards focus on other factors leading to the variability within abundance we see within our data.

#### Temperature Trends and Effects

```{r}
#| fig-cap: "Mean water temperature by year"

#FIGURE 6
#Calculaating annual water temp 
annual_water_temp <- combo_trawl %>%
  group_by(year) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE)
  )

#Figure 6
fig6 <- ggplot(annual_water_temp, aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  labs(
    x = "Year",
    y = "Mean Water Temperature"
  ) +
  theme_bw()

fig6

#savefigure 
ggsave("./figures/fig6.png", plot = fig6, width = 8, height = 6, dpi = 300)

```

Temperature is an incredibly important environmental factor that influences salmon survival, distribution, and health. Studies have shown that juveniles thrive within a very narrow temperature range as this is a very critical and metabolically demanding life stage. Variability in temperature can severely impact growth and stress.

From these long-term water temperature records, there seems to be a downward trend in overall temperature within the Delta (Fig. 6). However, management actions through the State Water Resources Control Board (95-1WR, May 1995) defined actions and implemented policy changes to "minimize losses of salmon and steelhead due to flow fluctuation." This led to an influx of cold water into the Delta, coinciding with several wet water years in a row. These actions led to a decrease in overall temperature of the Delta compared to previously warmer conditions.

```{r}
#| fig-cap: "Mean water temperature by year, 1995-2025"

#FIGURE 7
fig7 <- annual_water_temp %>% 
  filter(year > 1995) %>% 
  ggplot(aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  labs(
    x = "Year",
    y = "Mean Water Temperature"
  ) +
  theme_bw()

fig7

#savefigure 
ggsave("./figures/fig7.png", plot = fig7, width = 8, height = 6, dpi = 300)
```

After these policy changes, we have seen a steady increase in water temperatures in the Delta, with temperatures reaching over 19C in some years. This warming is significantly important for juvenile salmonid rearing in this habitat who need cooler waters to survive and grow.

```{r}
#| fig-cap: "Delta mean monthly water temperature"

#Figure 8
fig8 <- ggplot(salmonid_trawl, aes(x = month, y = WaterTemp)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  labs(
    x = "Month",
    y = "Water Temperature"
  ) +
  theme_bw()

fig8

#savefigure 
ggsave("./figures/fig8.png", plot = fig8, width = 8, height = 6, dpi = 300)
```

The Delta exhibits strong seasonal variability in water temperature, driven by changes in air temperature, inflows, and hydrologic conditions throughout the year (Fig. 8). During winter and early spring, temperatures tend to remain relatively cool as snowmelt and increased river flows enter the system. However, as summer progresses and flows decrease, water temperatures rise substantially across the Delta. This is incredibly important when considering juvenile fish migration timing and water policy changes that may aid in both juvenile outmigration survival, adult spawning, and egg incubation.

```{r}
#| fig-cap: "Water temperature (C) effects on juvenile salmon abundance"

#Figure 9
fig9 <- ggplot(salmonid_trawl, aes(x = WaterTemp, y = Count, color = Y)) +
  geom_point(alpha = 0.8) + 
  scale_color_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  labs(
    x = "Water Temperature (C)",
    y = "Daily Abundance Count"
  ) +
  theme_bw()

fig9

#savefigure 
ggsave("./figures/fig9.png", plot = fig9, width = 8, height = 6, dpi = 300)
```

Juvenile salmon abundance in the Delta appears to peak within an intermediate temperature range with some of the highest abundances recorded around 15C. This suggests that there is not only an optimal temperature in these moderate thermal conditions, but that there is also a temperature threshold juveniles can survive within. Temperatures on the lower end may reduce growth and metabolic rates, while the higher end may induce thermal stressors.

```{r}
#| include: false

#Model looking at the effects of flow on daily salmon abundance using a generalized additive model since non-linear effects 

m.temp <- gam(Count ~ s(WaterTemp),
              data = salmonid_trawl,
              family = nb())

saveRDS(m.temp, file = "./outputs/m.temp.rds")
```

```{r}
summary(m.temp)
```

In order to better understand the effects of temperature

```{r}
#| fig-cap: "Decadal shifts in Delta water temperature"

#FIGURE 10

#Changing the sample date column to recognize as date
combo_trawl$SampleDate <- as.Date(combo_trawl$SampleDate, format = "%m/%d/%Y")

#Get DOY for mapping
combo_trawl <- combo_trawl %>%
   mutate(DOY = yday(SampleDate))

#avg temp per day since multiple sampling per day
daily_water_temp <- combo_trawl %>%
  group_by(year, DOY) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

#Plotting water temp over time shifts
ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_line(color = "gray70", alpha = 0.5) +
  geom_line(
    data = subset(daily_water_temp, year >= 2020),
    aes(color = factor(year)),
    size = 1
  ) +
  theme_minimal()

ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_line(aes(color = year, linewidth = year, alpha = year)) +
  scale_color_viridis_c(option = "mako", direction = -1) +
  scale_linewidth(range = c(0.2, 1), guide = "none") +
  scale_alpha(range = c(1, 0.8), guide = "none") +
  theme_minimal()

ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_line(aes(color = year),
              method = "gam", se = FALSE, span = 0.25) +
  scale_color_viridis_c(option = "mako", direction = 1) +
  #scale_linewidth(range = c(0.3, 1.4), guide = "none") +
  #scale_alpha(range = c(0.3, 1), guide = "none") +
  theme_minimal()
```

Discuss decadal shifts over time– do an analysis to see if decades are sig different from each other.

#### Phenological Shifts

The timing of juvenile Chinook outmigration differs among run types, but is also affected by environmental variables, particularly flow and temperature. As a result, the year-to-year differences in *when* these fish pass through the Delta is likely a reflection of the conditions that they experience. We explored the beach seine data from the Delta Juvenile Fish Monitoring Program with the objective of visualizing patterns in the phenology of juvenile outmigration. We limited the data we used to samples from October through December to match the seasonality of the FMWT sampling (see above), and we used data from the sites that appeared to be sampled most consistently from 1977 through 2025. We pooled all run types together in this analysis: The data do include run type categories based on LAD (length-at-date), but these designations are notoriously inaccurate and, because of the limited months during which the data we used were collected, the salmon included in these data were almost certainly Spring- Late Fall- and Winter-Run fish with few Fall-Run Chinook among them. We tracked these counts over time by "day of water year", where day 0 was October 1 and day 365 (or 366) was September 30.

##### Accumulation Curves

To compare yearly migration patterns, we calculated the cumulative totals for each year, summing the counts for juvenile Chinook across all of the beach seine sites. We standardized these counts as a cumulative percentage of the annual total, and generated accumulation curves for each year.

The figure below shows the year-to-year differences in the timing and the pace at which outmigrating juvenile salmon enter the Delta. The earliest outmigrants appear in the beach seine samples around day 25 of the water year and 10% of the outmigrants have appeared by about day 100. Note that the rightmost accumulation curve is from the first year of the beach seine surveys, 1977. As an extreme outlier, the probability of some kind of sampling or data quality issue is quite high, and we have excluded these data from analyses until we learn more about the circumstances and the potential sources for error or some other explanatory factor.

```{r}
#| fig-cap: "Juvenile Chinook Accumulation Curves"

#FIGURE 11

#Load in Data
beachsieneALL <- read_csv(here("data", "processed", "seine_76_25.csv"))

#Preparing Data for Plot

## fish codes ----
# pull all IEP fish codes out to help with filtering 
# for fish data only (ie no inverts)
fish_codes <- 
    taxonomy %>% 
    filter(Phylum == "Chordata",
           # remove records where Species is NA
           # but retains eg "Bass unidentified"
           Species != "<NA>") %>%
    distinct(IEPFishCode)

# tidy data ----
temp <-
  beachsieneALL %>% 
  # select fishes only (includes empty hauls--seines w no fishes of any spp)
  filter(IEPFishCode %in% fish_codes$IEPFishCode) %>% 
  # provide columns w good names
  clean_names() %>% 
  # fix date & time variables, originally as chr
  mutate(date = mdy(sample_date), 
         time = times(sample_time), 
         .keep = "unused") %>% 
  # reorder columns
  relocate(date:time, .before = method_code)

## site selection -----
### effort -----
# calc the mean number of hauls at each site
# arbitrarily select only those sites whose mean equaled or exceeded 18

mean_effort <-
  temp %>% 
  group_by(location, region_code, station_code, year = year(date)) %>% 
  summarise(seine_hauls = n_distinct(station_code, date)) %>% 
  group_by(location, region_code, station_code) %>% 
  summarise(mean = mean(seine_hauls))

### reduce ----
# identify sites to be retained
selected_sites <-
  mean_effort %>% 
  filter(mean >= 18) %>% 
  left_join(sample_sites, 
            join_by(station_code == StationCode),
            # only keys (sites) from mean_effort are retained
            keep = FALSE,
            # drop unmatched dt5 rows
            unmatched = "drop",
            # expects each row in mean_effort matches at most 1 row in dt5
            relationship = "many-to-one") %>% 
  dplyr::select(-c(MethodCode, Location)) %>% 
  relocate(Latitude:Longitude, .after = region_code) %>% 
  # location is NOT unique; retained in case helps with orientation
  clean_names()

# select data from those minimum effort sites
seine <-
  temp %>% 
  filter(station_code %in% selected_sites$station_code)

## wy chinook -----
# add water year and day of water year
# select Chinook counts (no need for 'no catch'), begining with WY 1976 
# (Oct 1 1976 - Sep 30 1977) and ending with WY 2024 (Oct 1 2024 - end of 
# data set (July 31, close enough to the proper end of the WY for our purposes))

chisal <-
  seine %>% 
  # select only chinook and (mostly) complete water years
  filter(common_name == "Chinook Salmon" & 
           date >= "1976-10-01") %>% 
  # add water year (wy) and day of water year (wd) variables
  mutate(
    # water year
    wy = calcWaterYear(date),
    # day of water year (wd), wyMonth=10L bc water year begins Oct 1
    wd = get_waterYearDay(date, wyMonth = 10L)) %>% 
  # reorg & select columns
  relocate(wy:wd, .after = date) %>% 
  dplyr::select(c(location:water_temp, fork_length, race_by_length, count))

### daily counts -----
# aggregate to daily totals bc freq >1 seine per day
daily <- 
  chisal %>%
  group_by(wy, wd) %>%
  summarise(
    daily_total = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

### cumulative percent -----
# calc cumulative percentage for each wy
daily_cum <- 
  daily %>%
  group_by(wy) %>%
  arrange(wd) %>%
  mutate(
    annual_total = sum(daily_total),
    cum_total = cumsum(daily_total),
    cum_percent = 100 * cum_total / annual_total
  ) %>% 
  ungroup() %>% 
  arrange(wy, wd)

# outlier is 1977

# changed the colors where the older years are lighter and newer years are darker to see the phenological shifts in timing 
ggplot(daily_cum,
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako", direction = -1) +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Accumulation Curves",
       color = "water year") +
  theme_bw()


```

##### Thresholds

To examine the year-to-year differences more closely, we set 10%, 50% and 90% accumulation thresholds ("day_10", "day_50", and "day_90") for the migration timing, and compared these values across years.

```{r}
#| fig-cap: "Day of the Water Year for each Threshold (10, 50, 90%)"

#FIGURE 12

## ID thresholds ----
# day of water year when the 10%, 50% and 90% of total annual numbers are reached

# using 10%, 50%, and 90% accumulation thresholds (arbitrary)
thresholds <- c(10, 50, 90)

# gen df of water year days when threshold reached for each year
percentile_days <- 
  daily_cum %>%
  group_by(wy) %>%
  summarise(
    day_10 = min(wd[cum_percent >= 10]),
    day_50 = min(wd[cum_percent >= 50]),
    day_90 = min(wd[cum_percent >= 90]),
    .groups = "drop"
  )

# pivot data to long format to facilitate graphics
pctl_days <-
  percentile_days %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

# same data without 1997
pctl_days_red <-
  pctl_days %>% 
  filter(wy != "1977")

# figure shows the day of the water year for each threshold (10, 50, 90%) for each year
pctl_days %>% 
  ggplot(
    # color points according to threshold (10, 50, 90%)
    aes(x = wy, y = wd, color = percentile)
  ) +
  geom_point(size = 4) +
  scale_color_fish(discrete = TRUE,
                  option = "Hypsypops_rubicundus",
                  alpha = 0.6) +
  geom_smooth(data = pctl_days_red,
              alpha = 0.2,
              # show linear trends
              method = "lm",
              se = TRUE) +
  labs(title = "Chinook Salmon Outmigration Timing",
       subtitle = "1977 excluded from trend analyses",
       x = "Water Year",
       y = "Day of the Water Year",
       color = "Percentile") +
  theme_bw()
```

The figure above shows that, over time (1987-2025), all three thresholds are reached earlier and earlier. To explore this relationship, we modeled the timing for each benchmark (10, 50 and 90% of each year's total outmigrants) as a function of the water year.

```{r}
#| include: false

# reaching the 10% benchmark
m.day10.wy <- 
  brm(data = percentile_days, # use the percentile_days data
      # use a gaussian (normal) distribution
      family= gaussian,
      # specify the model
      day_10 ~ 1 + wy,
      # specify parameters for executing the Markov chains
      iter = 2000, warmup= 1000, chains= 4, cores=4,
      # setting the "seed" determines which randomnumbers will get sampled
      # so that team gets the same results when running the model
      seed = 4)

# reaching the 50% benchmark
m.day50.wy <- 
  brm(data = percentile_days, 
      # Choose a gaussian (normal) distribution
      family= gaussian,
      day_50 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 4)

# reaching the 90% benchmark
m.day90.wy <- 
  brm(data = percentile_days, 
      family= gaussian,
      day_90 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores =4,
      seed = 4)
```

Model summaries:

```{r}
summary(m.day10.wy)
summary(m.day50.wy)
summary(m.day90.wy)
```

To see how well these models performed and to determine the probability of a temporal effect, we plotted model performance and calculated the probabilities that the slope for each linear model differed from 0:

```{r}
#| fig-cap: "Model Plots for Year-Effect on 10, 50 & 90% Thresholds"

#FIGURE 13

# run the model plots for m.d_10.wy, m.d_50.wy, m.d_90.wy
plot(m.day10.wy)
plot(m.day50.wy)
plot(m.day90.wy) # less certain than the others

# calc the prob that the slope is less than 0
m.day10.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())
# virtually certain!

m.day50.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())

m.day90.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())
```

##### Water Year Type

We attempted to determine if there was an effect of water year type on the timing of Chinook outmigration. Environmental conditions vary from one year to the next, and a widely used metric for categorizing water availability for a water year is the Water Year Type: Each year, the California Department of Water Resources determines a water year type, a category based on the flow, unaltered by anthropogenic action, for the Sacramento River watershed. A year may be Critical, Dry, Below Normal, Above Normal, or Wet. We tested a model that included both Water Year and Water Year *Type* as explanatory factors:

```{r}
#| include: false

## year type ----
# first, add water year type to the percentile_days df
wy_type <- 
  read_excel(here("data", "processed", "cdec-water-year-type-jun-2025.xlsx")) %>% 
  mutate(wy = WY,
         wyt = factor(WYT),
         watershed = factor(Area),
         .keep = "unused")

# note that I'm using the df withOUT 1977
pctl_days_wyt <-
  wy_type %>% 
  filter(watershed == "Sacramento Valley") %>% 
  right_join(pctl_days_red,
             by = "wy") %>% 
  dplyr::select(-watershed) %>% 
  # specify order of the wyt levels
  mutate(wyt = fct_relevel(wyt, "C", "D", "BN", "AN", "W"),
         # center wy around the start of our data (1978)
         wy_centered = wy - 1978) %>% 
  relocate(wy_centered)

# run wy + wyt model with the centered variable, 1977 not included
m2.day10.wy.wyt <- 
  brm(data = pctl_days_wyt,
      family = gaussian,
      wd ~ 0 + wy_centered + wyt, 
      iter = 10000, warmup = 1000, chains = 4, cores = 4, seed = 42,
      # save fitted model (this one w the wy centered)
      file = here("data", "processed", "m2.day10.wy.wyt")
      )

```

Model summary and performance:

```{r}
summary(m2.day10.wy.wyt) 
# fixed Rhat issue but the estimates for different year types are all over the place
# and don't follow an ordered process and besides they all overlap considerably; the 
# dry year types (C & D) are the earliest, but wet is the next soonest! I still wonder
# if there is an interaction effect...

plot(m2.day10.wy.wyt)
```

```{r}
#| fig-cap: "Does Water Year Type Affect Outmigration Timing?"

#FIGURE 14

pctl_days_wyt %>% 
  filter(percentile == "day_10") %>% 
  ggplot(aes(x = wyt, y = wd, group = wyt)) +
  geom_boxplot() +
  labs(title = "Timing of 10% Annual Accumulation Threshold",
       x = "Water Year Type",
       y = "Day of the Water Year") +
  theme_bw()
```

##### Timing Shifts

The accumulation curves and the apparent temporal trends in the accumulation thresholds–the initial 10% of each year's salmon come 0.44 days earlier each year–suggest that some broad scale factor or suite of factors is affecting migratory phenology. Surprisingly, there does not seem to be an effect of water year type: Based on analysis of the timing of the 10% threshold, the measure is reach earliest in Critical and Dry years, but the third earliest timing occurs during Wet years. The distribution of these posteriors, however, all overlap considerably and it does seem reasonable to suggest that more direct effects are operating than these basin-wide patterns.

### Discussion and Suggested Action

How can these results be used in management decisions and policy changes...

A major limitation to our study is the fact that sizable (millions) but unknown numbers of hatchery fry are released into the system every year. Furthermore, the timing and the location of these releases have largely gone unrecorded; there is no way to know when or where these fish were released and even attempt to avoid or reduced the chances that they were included in these data. California's Central Valley Constant Fractional Marking program, whereby roughly 25% of these hatchery fish were adipose-clipped and tagged with coded wire tags, was begun in 2007, but the numbers are inconsistent and unreported.

Another challenge is the fact that genetic techniques are required to reliably distinguish between run types. Our data set includes run assignments based on length-at-date (LAD), but these are notoriously unreliable. Our focus on juveniles, captured during the fall and early winter, increases the likelihood that these fish are not fall-run and instead belong to one of the other four run types. It also increases the likelihood that these are wild-spawned and not hatchery-produced.
