---
title: "Long term water quality effects on juvenile salmon abundance in the Sacramento San-Joaquin River Delta"
author: "Peter Nelson, Ambrosio Rivera, Abigail Ward, Walt Vance"
format:
 pdf: default
editor: visual
execute:
  echo: false
  warning: false
  message: false
  results: "hide"
  dir: project
  fig-align: center
  fig-width: 8 
  fig-height: 6
---

## State of Juvenile Salmon in the Delta Report

Water is an important and valuable resource throughout the state of California. The Sacramento San-Joaquin River Delta is one of the most utilized water resources, providing water to nearly 27 million people across the state as well as supporting a \$1 billion/ year agricultural industry. Yet this area also contains a diversity of over 700 plant and animal species, many of which are important in recreational and community use. Understanding how to manage this resource in a way that both supports human endeavors and the vital ecosystem within this area is extremely important as we continue to face more extreme weather events.

### Salmonid Importance

Pacific salmon and steelhead are among the most ecologically, economically, and culturally significant species connected to the Sacramento–San Joaquin River Delta. The California salmon industry is valued at \$1.4 billion, contributing to nearly 23,000 jobs annually. This economy depends on healthy runs of Chinook, coho, and steelhead utilizing these waterways. The Delta plays a vital role in the life cycle of these salmonids. Around two thirds of species of California's adult Pacific salmon pass through the Delta on their way upstream to spawn, while juveniles utilize the area as rearing habitat. This transitional zone provides an abundance of food, refuge and ideal conditions for physiological adaptations as they prepare for the marine environment.

However, salmonid populations face pressure from changes to flow regimes, water temperatures, habitat loss, and extreme weather events. Reduced snow melt runoff, prolonged drought, and flood events have disrupted migration timing and reproductive success. Effective Delta management must then balance water supply with environmental needs, habitat restoration, and temperature control measures to sustain both salmon populations and the communities that depend on them.

### Report Objectives

This assessment aims to evaluate the current state of juvenile salmonid abundances in the Sacramento–San Joaquin River Delta and how these abundances have changed over time as an effect of ecological parameters linked to water management practices. This report focuses specifically on key species such as steelhead and Chinook salmon which rely on the Delta as critical migratory and rearing habitat.

To conduct our analyses, we utilize long-term monitoring data from the Delta Juvenile Fish Monitoring Program (DJFMP). This dataset provides multi-decadal information on juvenile salmonid abundance, allowing for evaluation of temporal trends and potential relationships with flow, temperature, and other ecological parameters. Through this, the study seeks to better understand how water management strategies may influence salmonid population dynamics within the Delta and propose how water resource managers may better work in the Delta.

### Results

#### Salmonid Abundance

```{r}
#Load packages
library(here)
library(readr)
library(tidyverse)
library(readxl) 
library(dplyr)
library(janitor) 
library (stringr)
library(openxlsx)
library(stats)
library(car)
library(ggplot2)
library(ggpubr)
library(lubridate)
library(MASS)
library(brms)
library(chron)
library(gridExtra)
library(dataRetrieval)
remotes::install_github("doi-usgs/EflowStats@v5.2.0")
library(EflowStats)
```

```{r}
#Load in data 
sample_sites <- read_csv(here("data", "processed", "sample_sites.csv"))
taxonomy <- read_csv(here("data", "processed", "taxonomy.csv"))
trawl_02_25 <- read_csv(here("data", "processed", "trawl_02_25.csv"))
trawl_76_01 <- read_csv(here("data", "processed", "trawl_76_01.csv"))

#Separate sample date into day, month year
trawl_76_01 <- trawl_76_01 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_02_25 <- trawl_02_25 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)


trawl_76_01 <- trawl_76_01 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

trawl_02_25 <- trawl_02_25 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

#Create combination of datasets
combo_trawl <- bind_rows(trawl_76_01, trawl_02_25)
combo_trawl$year <- as.numeric(combo_trawl$year)

#Create season column
combo_trawl <- combo_trawl %>%
  mutate(
    season = case_when(
      month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      month %in% c("Mar", "Apr", "May") ~ "Spring",
      month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      month %in% c("Sep", "Oct", "Nov") ~ "Fall"
    )
  )

#Salmonid only dataset
#Filtering for salmonids and for specific sites  
salmonid_trawl <- combo_trawl %>% 
  filter(OrganismCode == "CHN" | OrganismCode == "RBT") 

#Summming the counts per year to get abundances and getting runs 
salmonid_trawl <- salmonid_trawl %>% 
  mutate(
    run = case_when(
      OrganismCode == "RBT" ~ "steelhead",
      !is.na(RaceByLength) ~ RaceByLength,
      !is.na(RaceByTag) ~ RaceByTag,
      TRUE ~ NA_character_
    )
  )


```

```{r}
#| fig-cap: "Juvenile salmon abundance over time"

ggplot(
  salmonid_trawl %>%
    group_by(year) %>%
    summarise(annual_abundance = sum(Count, na.rm = TRUE),
              .groups = "drop"),
  aes(x = year,
      y = annual_abundance)
) +
  geom_line(color = "black") +
  geom_smooth(method = "lm")+
  geom_point() +
  labs(
    x = "Year",
    y = "Total Annual Salmonid Abundance",
    title = "Annual Salmonid Abundance Over Time"
  ) +
  theme_bw()
```

```{r}
#| include: false

#Add annual abundance of fish column
salmonid_trawl <- salmonid_trawl %>%
  group_by(year) %>%
  mutate(annual_abundance = sum(Count, na.rm = TRUE)) %>%
  ungroup()

#Create annual abundace dataset
annual_data <- salmonid_trawl %>%
  group_by(year) %>%
  summarise(
    annual_abundance = sum(Count, na.rm = TRUE),
    .groups = "drop"
  )

#Perform a negative binomial glm to look at annual abundace over the years
model_nb <- glm.nb(annual_abundance ~ year,
                   data = annual_data)

summary(model_nb)


```

```{r}
#| include: false
exp(-0.01571)
```

```{r}
#| include: false
100*(1-0.9844)
```

Overall, we've seen a 1.56% decline in juvenile salmon abundances since 1976. Although there have been specific peaks throughout time, this general trend in the Delta is extremely concerning as a downward trend in juvenile abundance can give insight into how changes to climate and habitat have had long term negative effects. Because juvenile survival strongly influences adult return rates, sustained declines at this life stage can translate into reduced numbers of spawning adults in subsequent years. This trend is vital to understand as we begin to explore how water management strategies can uplift and boost these populations.

#### Flow Rate Trends and Effects

```{r}
#Load in flow data
flow_70_83 <- read_csv(here("data", "processed", "dayflow-results-1970-1983.csv"))
flow_84_96 <- read_csv(here("data", "processed", "dayflow-results-1984-1996.csv"))
flow_97_23 <- read_csv(here("data", "processed", "dayflow-results-1997-2023.csv"))
water_year <- read_xlsx(here("data", "processed", "cdec-water-year-type-jun-2025.xlsx"))

#Filter Water Year Data
water_year <- water_year %>%
  rename(Year = WY) %>% 
  filter(Area == "Sacramento Valley")

#Combine Flow Data
allflow <- bind_rows(flow_70_83, flow_84_96, flow_97_23)
allflow <- allflow %>% 
 rename(SampleDate = Date) 

#Correct Flow Dates
allflow <- allflow %>%
  mutate(SampleDate = as.Date(SampleDate, format = "%m/%d/%Y"))


#Get water year for all NAs
allflow <- allflow %>%
  left_join(
    water_year %>% dplyr::select(Year, WYT),
    by = "Year"
  ) %>%
  mutate(
    Y = WYT  
  ) %>%
  dplyr::select(-WYT) 

#Caluclate daily abundace 
daily_salmonid_abundance <- salmonid_trawl %>%
  group_by(SampleDate, CommonName, run) %>%
  summarise(
    total_fish = sum(Count, na.rm = TRUE),
    .groups = "drop"
  ) 


#Merge abundances back in 
salmonid_trawl <- salmonid_trawl %>%
  left_join(daily_salmonid_abundance,
            by = c("SampleDate", "CommonName", "run"))

#Correct Flow Dates
salmonid_trawl <- salmonid_trawl %>%
  mutate(SampleDate = as.Date(SampleDate, format = "%m/%d/%Y"))


#Merge with flow data
salmonid_trawl <- salmonid_trawl %>%
  left_join(
    allflow %>% dplyr::select(SampleDate, OUT, Y, Year),
    by = "SampleDate"
  )

 
```

```{r}
#| fig-cap: "Mean annual flow by water year type"

#annual flow for Delta
annual_flow <- allflow %>%
  group_by(Year, Y) %>%                               
  summarise(
    mean_OUT = mean(OUT, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(annual_flow, aes(x = Year, y = mean_OUT, fill = Y)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  scale_x_continuous(
    breaks = seq(min(annual_flow$Year),
                 max(annual_flow$Year),
                 by = 5)
  ) +
  labs(
    x = "Year",
    y = "Mean Annual Flow (cfs)"
  ) +
  theme_bw()

```

```{r}
#| fig-cap: "Linear trend of annual flow"

ggplot(annual_flow, aes(x = Year, y = mean_OUT)) +
  geom_line(alpha = 0.3) +           
  geom_smooth(method = "lm") + 
  labs(
    x = "Year",
    y = "Mean Annual Flow (cfs)"
  ) +
  theme_bw()
```

Over the past several decades, we've seen increasingly drier conditions and a reduction in overall water availability. These conditions have severe consequences for juvenile salmonids who rely on freshwater flows through the Delta. This long-term monitoring has suggested that average flows through the Delta have trended downward over time (Fig.3). We also have see drier conditions in the Sacramento River System as a whole, with an increase in "Dry" and "Critical" years since 2000 (Fig. 2).

However, even in years classified as "Above Normal", the amount of water flowing out of the Delta can remain relatively low due to upstream storage, diversions, and exports to Southern California, agriculture and the Bay area (Fig. 2). This means that the ecological benefits of wetter years are not always fully realized within the Delta ecosystem or by the species within. Water management decisions during these times, then, become critical to sustaining populations and shaping the ecological conditions of the Delta.

```{r}
#| fig-cap: "Daily salmonid abundance as a function of flow"

ggplot(salmonid_trawl, aes(x = OUT, y = Count, color = Y)) +
  geom_point()  +
  scale_color_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  labs(
    x = "Daily Flow (cfs)",
    y = "Daily Abundance Count"
  ) +
  theme_bw()

```

```{r}
#| include: false

#Model looking at the effects of flow on daily salmon abundance 
m.flow <- brm(
  Count ~ log(OUT + 1),
  family = negbinomial(),
  data = salmonid_trawl,
  chains = 4, cores = 4, iter = 4000,
  algorithm = 'fullrank'
)
```

```{r}
summary(m.flow)
```

Discuss here about how daily flows and water years effect abundance, discuss the model, etc...

#### Temperature Trends and Effects

```{r}
#| results: asis
#| fig-cap: "Mean water temperature by year"

#Calculaating annual water temp 
annual_water_temp <- combo_trawl %>%
  group_by(year) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE)
  )

ggplot(annual_water_temp, aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  labs(
    x = "Year",
    y = "Mean Water Temperature"
  ) +
  theme_bw()
```

From these long-term water temperature records, there seems to be a downward trend in overall temperature within the Delta. However, management actions through the State Water Resources Control Board (95-1WR, May 1995) defined actions and implemented policy changes to "minimize losses of salmon and steelhead due to flow fluctuation." This led to an influx of cold water into the Delta, coinciding with several wet water years in a row. These actions led to a decrease in overall temperature of the Delta compared to previously warmer conditions.

```{r}
#| results: asis
#| fig-cap: "Mean water temperature by year, 1995-2025"

annual_water_temp %>% 
  filter(year > 1995) %>% 
  ggplot(aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  labs(
    x = "Year",
    y = "Mean Water Temperature"
  ) +
  theme_bw()
```

Now, however, we are beginning to trend back upward to warming temperatures. Recent years have seen particularly high peaks in mean Delta water temperatures, reaching above 18 degrees Celcius.

```{r}
#| results: asis
#| fig-cap: "Delta mean monthly water temperature"

ggplot(salmonid_trawl, aes(x = month, y = WaterTemp)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  labs(
    x = "Month",
    y = "Water Temperature"
  ) +
  theme_bw()
```

The Delta exhibits strong seasonal variability in water temperature, driven by changes in air temperature, inflows, and hydrologic conditions throughout the year. During winter and early spring, temperatures tend to remain relatively cool as snowmelt and increased river flows enter the system. However, as summer progresses and flows decrease, water temperatures rise substantially across the Delta. This is incredibly important when considering juvenile fish migrations and the

```{r}
#| results: asis
#| fig-cap: "Water temperature (C) effects on juvenile salmon abundance"

ggplot(salmonid_trawl, aes(x = WaterTemp, y = Count, color = Y)) +
  geom_point(alpha = 0.8) + 
  scale_color_manual(values = c("W"= "#021893", "AN"= "#3537ae", "BN" = "#740699" , "D" = "#b70b0b", "C" = "#990623" ), labels = c ("W" = "Wet", "AN" = "Above Normal", "BN" = "Below Normal", "D"= "Dry", "C"= "Critcal"), name = "Water Year Type") + 
  labs(
    x = "Water Temperature (C)",
    y = "Daily Abundance Count"
  ) +
  theme_bw()
```

Something about optimal water temp for juveniles and the effects of water temp even in dry years can lead to high abundance

```{r}
#| fig-cap: "Decadal shifts in Delta water temperature"

#Changing the sample date column to recognize as date
combo_trawl$SampleDate <- as.Date(combo_trawl$SampleDate, format = "%m/%d/%Y")

#Get DOY for mapping
combo_trawl <- combo_trawl %>%
   mutate(DOY = yday(SampleDate))

#avg temp per day since multiple sampling per day
daily_water_temp <- combo_trawl %>%
  group_by(year, DOY) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

#Plot
ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_line(color = "gray70", alpha = 0.5) +
  geom_line(
    data = subset(daily_water_temp, year >= 2020),
    aes(color = factor(year)),
    size = 1
  ) +
  theme_minimal()

ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_line(aes(color = year, linewidth = year, alpha = year)) +
  scale_color_viridis_c(option = "mako", direction = -1) +
  scale_linewidth(range = c(0.2, 1), guide = "none") +
  scale_alpha(range = c(1, 0.8), guide = "none") +
  theme_minimal()

ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_smooth(aes(color = year),
              method = "loess", se = FALSE, span = 0.25) +
  scale_color_viridis_c(option = "rocket", direction = 1) +
  #scale_linewidth(range = c(0.3, 1.4), guide = "none") +
  #scale_alpha(range = c(0.3, 1), guide = "none") +
  theme_minimal()
```

Discuss decadal shifts over time– do an analysis to see if decades are sig different from each other.

#### Phenological Shifts

```{r}
#| fig-cap: "Shifts in juvenile outmigration time "

#Load in Data
beachsieneALL <- read_csv(here("data", "processed", "seine_76_25.csv"))

#Preparing Data for Plot

## fish codes ----
# pull all IEP fish codes out to help with filtering 
# for fish data only (ie no inverts)
fish_codes <- 
    taxonomy %>% 
    filter(Phylum == "Chordata",
           # remove records where Species is NA
           # but retains eg "Bass unidentified"
           Species != "<NA>") %>%
    distinct(IEPFishCode)

# tidy data ----
temp <-
  beachsieneALL %>% 
  # select fishes only (includes empty hauls--seines w no fishes of any spp)
  filter(IEPFishCode %in% fish_codes$IEPFishCode) %>% 
  # provide columns w good names
  clean_names() %>% 
  # fix date & time variables, originally as chr
  mutate(date = mdy(sample_date), 
         time = times(sample_time), 
         .keep = "unused") %>% 
  # reorder columns
  relocate(date:time, .before = method_code)

## site selection -----
### effort -----
# calc the mean number of hauls at each site
# arbitrarily select only those sites whose mean equaled or exceeded 18

mean_effort <-
  temp %>% 
  group_by(location, region_code, station_code, year = year(date)) %>% 
  summarise(seine_hauls = n_distinct(station_code, date)) %>% 
  group_by(location, region_code, station_code) %>% 
  summarise(mean = mean(seine_hauls))

### reduce ----
# identify sites to be retained
selected_sites <-
  mean_effort %>% 
  filter(mean >= 18) %>% 
  left_join(sample_sites, 
            join_by(station_code == StationCode),
            # only keys (sites) from mean_effort are retained
            keep = FALSE,
            # drop unmatched dt5 rows
            unmatched = "drop",
            # expects each row in mean_effort matches at most 1 row in dt5
            relationship = "many-to-one") %>% 
  dplyr::select(-c(MethodCode, Location)) %>% 
  relocate(Latitude:Longitude, .after = region_code) %>% 
  # location is NOT unique; retained in case helps with orientation
  clean_names()

# select data from those minimum effort sites
seine <-
  temp %>% 
  filter(station_code %in% selected_sites$station_code)

## wy chinook -----
# add water year and day of water year
# select Chinook counts (no need for 'no catch'), begining with WY 1976 
# (Oct 1 1976 - Sep 30 1977) and ending with WY 2024 (Oct 1 2024 - end of 
# data set (July 31, close enough to the proper end of the WY for our purposes))

chisal <-
  seine %>% 
  # select only chinook and (mostly) complete water years
  filter(common_name == "Chinook Salmon" & 
           date >= "1976-10-01") %>% 
  # add water year (wy) and day of water year (wd) variables
  mutate(
    # water year
    wy = calcWaterYear(date),
    # day of water year (wd), wyMonth=10L bc water year begins Oct 1
    wd = get_waterYearDay(date, wyMonth = 10L)) %>% 
  # reorg & select columns
  relocate(wy:wd, .after = date) %>% 
  dplyr::select(c(location:water_temp, fork_length, race_by_length, count))

### daily counts -----
# aggregate to daily totals bc freq >1 seine per day
daily <- 
  chisal %>%
  group_by(wy, wd) %>%
  summarise(
    daily_total = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

### cumulative percent -----
# calc cumulative percentage for each wy
daily_cum <- 
  daily %>%
  group_by(wy) %>%
  arrange(wd) %>%
  mutate(
    annual_total = sum(daily_total),
    cum_total = cumsum(daily_total),
    cum_percent = 100 * cum_total / annual_total
  ) %>% 
  ungroup() %>% 
  arrange(wy, wd)


# outlier is 1977
ggplot(daily_cum,
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = factor(wy))) +
  geom_line() +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Water Year Chinook Accumulation Curves") +
  theme_bw() +
  theme(legend.position = "none")

#changed the colors where the older years are lighter and newer years are darker to see the phenological shifts in timing 
ggplot(daily_cum,
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
 scale_color_viridis_c(option = "mako", direction = -1) +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Water Year Chinook Accumulation Curves") +
  theme_bw() +
  theme(legend.position = "none")


```

Peter choose one of the two figures above and write your description here!

### Discussion and Suggested Action

How can these results be used in management decisions and policy changes...
