---
title: "exploring seine data"
author: "Pete Nelson"
format: pdf
editor: visual
---

# beach seine data

Goals:

-   select appropriate stations (drop those sampled infrequently or outside area of interest)
-   add station location

```{r}

library(tidyverse)
library(ggplot2)

seine_long <- 
  read_rds("~/Documents/projects/ds4eeb/Pete_local_data/seine_long.rds")

seine_long_min_effort <-
  read_rds("~/Documents/projects/ds4eeb/Pete_local_data/seine_long_min_effort.rds")

seine <- 
  read_rds("~/Documents/projects/ds4eeb/Pete_local_data/seine.rds")

```

### fall salmon data

```{r}
fall_salmon <- 
  seine_long_min_effort %>% 
  mutate(year = year(date), month = month(date)) %>% 
  relocate(year:month, .before = date) %>% 
  filter(common_name == "Chinook Salmon" | common_name == "No catch") %>% 
  # months chose to correspond to FMWT
  filter(month >= 9 & month <= 12)

temp <-
  fall_salmon %>% 
  group_by(region_code, year) %>% 
  summarise(total = sum(count, na.rm = TRUE))

temp %>% 
  ggplot(aes(x = region_code, y = total)) +
  geom_point() +
  theme_bw()

temp <-
  fall_salmon %>% 
  group_by(station_code, year) %>% 
  summarise(total = sum(count, na.rm = TRUE))

temp %>% 
  ggplot(aes(x = station_code, y = total)) +
  geom_point() +
  theme_bw()

temp %>% 
  ggplot(aes(x = year, y = total)) +
  geom_point() +
  theme_bw()

# consider: calc monthly mean temperature and overlay this on the salmon totals
# accumulation curves
```

# analyses

The challenges of nailing the location information, plus mapping it effectively feel a bit much, given our deadlines.

I want to focus on analyses. Thinking that I'll generate Delta-wide accumulation curves for chinook, and look at the timing for 10%, 50% and 90% events as a function of Delta outflow. Because those events should be normally distributed, the stats should be straightforward (no weird distributions). If Delta outflow numbers prove hard to get, we could work with categorical water year type, sensu DWR. One good thing about water year type is that they directly reflect water usage, so we're looking at the possible effects of human action as well as environmental variability.

## accumulation curves

```{r}

# select Chinook counts (no need for 'no catch'), begining with WY 1976 (Oct 1 1976 - Sep 30 1977) and ending with WY 2024 (Oct 1 2024 - end of data set (July 31, close enough to the proper end of the WY for our purposes))

chisal <-
  seine_long_min_effort %>% 
  filter(common_name == "Chinook Salmon" & date >= "1976-10-01")
  
# add water year, including wy_day
library(dataRetrieval)
library(EflowStats)

chisal <-
  chisal %>% 
  mutate(
    # water year
    wy = calcWaterYear(date),
    # day of water year (wd), wyMonth=10L bc water year begins Oct 1
    wd = get_waterYearDay(date, wyMonth = 10L)) %>% 
  # reorg & select columns
  relocate(wy:wd, .after = date) %>% 
  select(c(location:water_temp, fork_length, race_by_length, count))

# aggregate to daily totals bc freq >1 seine per day
daily <- 
  chisal %>%
  group_by(wy, wd) %>%
  summarise(
    daily_total = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

# calc cumulative percentage for each wy
daily_cum <- 
  daily %>%
  group_by(wy) %>%
  arrange(wd) %>%
  mutate(
    annual_total = sum(daily_total),
    cum_total = cumsum(daily_total),
    cum_percent = 100 * cum_total / annual_total
  ) %>% 
  ungroup() %>% 
  arrange(wy, wd)

# plot accumulation curves
ggplot(daily_cum,
       aes(x = wd,
           y = cum_percent,
           group = wy)) +
  geom_line(alpha = 0.3) +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Water Year Chinook Accumulation Curves") +
  theme_bw()

# color by year
ggplot(daily_cum,
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = factor(wy))) +
  geom_line() +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Water Year Chinook Accumulation Curves") +
  theme_bw() +
  theme(legend.position = "none")

# outlier is 1977 (see below too)
```

The outlier, with 50% accumulate about wd 235, is 1977. Because it's the first year of the time series, I think we're justified dropping this from the analyses at least (or if) until we're able to learn more about sampling that year

## identify thresholds

```{r}
# day of water year when the 10%, 50% and 90% of total annual numbers are reached
thresholds <- c(10, 50, 90)

percentile_days <- 
  daily_cum %>%
  group_by(wy) %>%
  summarise(
    day_10 = min(wd[cum_percent >= 10]),
    day_50 = min(wd[cum_percent >= 50]),
    day_90 = min(wd[cum_percent >= 90]),
    .groups = "drop"
  )

percentile_days
write_csv(percentile_days, "~/Documents/projects/ds4eeb/final_project/data/percentile_days.csv")
```

Say something here

```{r}
ggplot(daily_cum,
       aes(wd, cum_percent, group = wy)) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  theme_bw()
```
