---
title: "walt_djfmp_explore"
format: html
---

## Tidying Data:

```{r}
#loading packages
library(ggplot2)
library(here)
library(tidyverse)
library(dplyr)

```

```{r}
#reading data
sample_sites <- read.csv(here("data","processed","sample_sites.csv"))
taxonomy <- read.csv(here("data","processed","taxonomy.csv"))
seine <- read.csv(here("data","processed","seine_76_25.csv"))
trawl_02_25 <- read.csv(here("data","processed","trawl_02_25.csv"))
trawl_76_01 <- read.csv(here("data","processed","trawl_76_01.csv"))
```

```{r}
#Separate sample date into day, month year (thanks abby for the code)
trawl_76_01 <- trawl_76_01 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_02_25 <- trawl_02_25 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

seine <- seine %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_76_01 <- trawl_76_01 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

trawl_02_25 <- trawl_02_25 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

seine <- seine %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )


combo_trawl <- bind_rows(trawl_76_01, trawl_02_25)
combo_trawl$year <- as.numeric(combo_trawl$year)

```

### Joining Tables for Taxonomy

```{r}
trawl_joined <- combo_trawl %>%
  left_join(taxonomy, by = "OrganismCode")

#trawl_counts_taxonomy <- trawl_joined %>% 
  #filter(CommonName != "No catch",Family != "NA") %>% 
  
  #group_by(Location, year, Family) %>% 
  #summarize(Count = sum(Count, na.rm = TRUE)) 
  
```

```{r}
#joing seine and toxonmy tables
seine_tax <- seine %>%
  left_join(taxonomy, by = "OrganismCode")

```

### Totaling Counts by location, year, and Species

```{r}
combo_trawl_counts <- combo_trawl %>% 
  group_by(Location, year, CommonName) %>% 
  summarize(Count = sum(Count, na.rm = TRUE)) %>% 
  filter(CommonName != "No catch")
```

```{r}
seine_counts <- seine_tax %>% 
  group_by(Location, year, OrganismCode) %>% 
  summarize(Count = sum(Count, na.rm = TRUE)) %>% 
  filter(OrganismCode != "NOFISH")
```

### Totaling Counts of Unique Species In Seine Data

```{r}
seine_richness <- seine_tax %>% 
  select(Location, year, OrganismCode, CommonName.x, Family, NonNative) %>% 
  group_by(Location, year, OrganismCode, NonNative) %>% 
  summarize(count = n()) %>% #creates a column that totals the columns of that loc, year, and organism
  filter(OrganismCode != "NOFISH") %>% 
  group_by(Location, year, NonNative) %>% 
  summarize(count = n()) # allows us to total the unqiue number of organisms in each location each year seperated by native status
```

### Graphing time!

```{r}
count_year <- ggplot(data = combo_trawl_counts, mapping = aes(x = year, y = Count)) +
  geom_point() +
  facet_wrap(Location ~ .) +
  ylim(0,250000)

count_year
```

```{r warning = FALSE, message = FALSE}
count_species <- ggplot(data = combo_trawl_counts, mapping = aes(x = year, y = Count)) +
  geom_point(aes(color= Location)) +
  geom_line(aes(color= Location)) +
  facet_wrap(CommonName ~ ., scales = "free_y")

count_species
```

```{r warning = FALSE, message = FALSE}
count_total <- ggplot(data = combo_trawl_counts, mapping = aes(x = year, y = Count)) +
  geom_point(aes(color=CommonName)) +
  geom_smooth(aes(color=CommonName)) +
  ylim(0,100000)

count_total
```

```{r warning = FALSE, message = FALSE}
count_family <- ggplot(data = trawl_counts_taxonomy, mapping = aes(x = year, y = Count)) +
  geom_point(aes(color= Location)) +
  geom_line(aes(color= Location)) +
  facet_wrap(Family ~ ., scales = "free_y")

count_family
```

### Stacked Area

```{r warning = FALSE, message = FALSE}
count_area <- ggplot(data = trawl_counts_taxonomy, mapping = aes(x = year, y = Count)) +
  geom_area(aes(fill= Family)) +
  facet_wrap(Location ~ ., scales = "free_y") +
  scale_y_log10()

count_area

```

```{r warning = FALSE, message = FALSE}
count_area2 <- ggplot(data = trawl_counts_taxonomy, mapping = aes(x = year, y = Count)) +
  geom_area(aes(fill= Family)) +
  scale_y_log10()

count_area2

```

## Native vs Invasive

```{r}
native_trawl <- trawl_joined %>% 
   filter(CommonName != "No catch",NonNative != "NA") %>%
  group_by(Location, year, Family, NonNative) %>% 
  summarize(Count = sum(Count, na.rm = TRUE)) 

```

```{r}
count_nonnative <- ggplot(data = native_trawl, mapping = aes(x = year, y = Count)) +
  geom_point(aes(color=NonNative)) +
  facet_wrap(Location ~ .) +
  ylim(0,250000)

count_nonnative
```

```{r warning = FALSE, message = FALSE}
native_area <- ggplot(data = native_trawl, mapping = aes(x = year, y = Count)) +
  geom_area(aes(fill= NonNative)) +
  facet_wrap(Location ~ .) +
  scale_y_log10()

native_area

```

## Seine Data Graphing

### native non native

```{r}
ggplot()
```

## Trawl Data on Focal Species

look at how anchovies salmonids chinnok coorelate over time by location and by temperature.

#### Organizing counts of focal salmonids per year by station

```{r}
#creating a filtered data table with counts of salmonids 
focal_trawl_counts <- combo_trawl %>% 
  filter(OrganismCode == "CHN"| OrganismCode ==  "CHO" | OrganismCode ==  "RBT") %>% 
  group_by(StationCode, year, OrganismCode) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")
  
```

```{r}
salmon_spp <- ggplot(data= focal_trawl_counts, aes(x = year, y = Count, fill = StationCode)) +
  geom_line(aes(color = StationCode)) +
  facet_wrap(OrganismCode ~., scales = "free_y")

salmon_spp

```

```{r}
salmon_spp <- ggplot(data= focal_trawl_counts, aes(x = year, y = Count )) +
  geom_line() +
  facet_wrap(OrganismCode ~., scales = "free_y")

salmon_spp
```

```{r}
salmon_loc <- ggplot(data= focal_trawl_counts, aes(x = year, y = Count, fill = OrganismCode)) +
  geom_line(aes(color = OrganismCode)) +
  facet_wrap(StationCode ~., scales = "free_y")

salmon_loc
```

#### Organizing counts of focal salmonids per year by station (filtering by stations with abundant data)

```{r}
focal_filter <- combo_trawl %>% 
  filter(OrganismCode == "CHN"| OrganismCode ==  "RBT") %>% 
  filter(StationCode == "SB018M" | 
        StationCode == "SB018N" |
        StationCode == "SB018S" | 
        StationCode == "SB018X" |
        StationCode == "SJ054M" |
        StationCode == "SR055M") %>% 
  group_by(StationCode, year, OrganismCode) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")
  
```

```{r}
salmon_spp_station <- ggplot(data= focal_filter, aes(x = year, y = Count, fill = StationCode)) +
  geom_line(aes(color = StationCode)) +
  facet_wrap(OrganismCode ~., scales = "free_y")

salmon_spp_station

```

```{r}
salmon_spp <- ggplot(data= focal_filter, aes(x = year, y = Count )) +
  geom_line() +
  facet_wrap(OrganismCode ~., scales = "free_y")

salmon_spp
```

```{r}
salmon_loc_2 <- ggplot(data= focal_trawl_counts, aes(x = year, y = Count, fill = OrganismCode)) +
  geom_line(aes(color = OrganismCode)) +
  facet_wrap(StationCode ~., scales = "free_y")

salmon_loc_2
```

#### Including watertemp?

```{r}
focal_filter2 <- combo_trawl %>% 
  filter(OrganismCode == "CHN"| OrganismCode ==  "RBT") %>% 
  filter(StationCode == "SB018M" | 
        StationCode == "SB018N" |
        StationCode == "SB018S" | 
        StationCode == "SB018X" |
        StationCode == "SJ054M" |
        StationCode == "SR055M") %>% 
  group_by(year) %>% 
  mutate(avg_temp = mean(WaterTemp, na.rm = TRUE), .groups = "drop") %>% 
  group_by(StationCode, year, OrganismCode, avg_temp) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")
```

```{r}
#plot for temperature on salmon populations through time
salmon_spp_station <- ggplot(data= focal_filter2, aes(x = year, y = Count, color = avg_temp)) +
  geom_line() +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap(OrganismCode ~., scales = "free_y") +
  theme_classic()
  


salmon_spp_station
```

```{r}
salmon_spp_station2 <- ggplot(data= focal_filter2, aes(x = avg_temp, y = Count, color = year)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap(OrganismCode ~., scales = "free_y") +
  theme_classic()
salmon_spp_station2
```

#### now for flow

```{r}
focal_filter_flow <- combo_trawl %>% 
  filter(OrganismCode == "CHN"| OrganismCode ==  "RBT") %>% 
  filter(StationCode == "SB018M" | 
        StationCode == "SB018N" |
        StationCode == "SB018S" | 
        StationCode == "SB018X" |
        StationCode == "SJ054M" |
        StationCode == "SR055M") %>% 
  group_by(year) %>% 
  mutate(avg_flow = mean(FlowmeterDifference, na.rm = TRUE), .groups = "drop") %>% 
  group_by(StationCode, year, OrganismCode, avg_flow) %>% 
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")
```

```{r}
salmon_spp_station_flow <- ggplot(data= focal_filter_flow, aes(x = year, y = Count, color = avg_flow)) +
  geom_line() +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap(OrganismCode ~., scales = "free_y")


salmon_spp_station_flow
```

```         
```

### salmonids over years (thanks for the base code abbie!)

```{r}
salmonids_by_year <- combo_trawl %>%
  filter(!is.na(CommonName)) %>%
  filter(CommonName == "Chinook Salmon" | CommonName == "steelhead") %>% 
  group_by(year, CommonName) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(year, desc(count))

salmonids_by_year

#Chinook by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="Chinook Salmon"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

#Steelhead by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="steelhead"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

```

### looking at avg water flow by year (adjusted for flow from temp code written by abbie)

```{r}
#Cal avg water temp by year
avg_flow_by_year <- combo_trawl %>%
  group_by(year) %>%
  summarise(avg_flow = mean(FlowmeterDifference, na.rm = TRUE),
            .groups = "drop")

#Filter for chinook & steelhead individually
ch_by_year <- salmonids_by_year %>% 
  filter(CommonName == "Chinook Salmon")
sh_by_year <- salmonids_by_year %>% 
  filter(CommonName == "steelhead")

#flow/ salmon plot data
flow_ch_data <- left_join(ch_by_year, avg_flow_by_year, by = "year")
flow_sh_data <- left_join(sh_by_year, avg_flow_by_year, by = "year")

#Scale for dual axis plotting chinook
scale_factor_ch <- max(flow_ch_data$count, na.rm = TRUE) / 
                max(flow_ch_data$avg_flow, na.rm = TRUE)
flow_ch_data <- flow_ch_data %>%
  mutate(flow_scaled = avg_flow * scale_factor_ch)

#Scale for dual axis plotting steelhead
scale_factor_sh <- max(flow_sh_data$count, na.rm = TRUE) / 
                max(flow_sh_data$avg_flow, na.rm = TRUE)
flow_sh_data <- flow_sh_data %>%
  mutate(flow_scaled = avg_flow * scale_factor_sh)


#Dual axis plot chinook 
ggplot(flow_ch_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = flow_scaled, color = "Avg Water Flow"), 
            linetype = "solid", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor_ch,
                        name = "Average Water Flow")
  ) +
  
  theme_minimal() 

#Dual axis plot steelhead 
ggplot(flow_sh_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = flow_scaled, color = "Avg Water Flow"), 
            linetype = "solid", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor_sh,
                        name = "Average Water Flow")
  ) +
  
  theme_minimal() 


```
