---
title: "Ward_DJFMP_explore"
author: "Ward, A."
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls()) #clean environment

library(here)
knitr::opts_knit$set(
  root.dir = here::here()
)

knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r packages, include=FALSE}
library(readr)
library(tidyverse)
library(readxl) 
library(dplyr)
library(janitor) 
library (stringr)
library(openxlsx)
library(stats)
library(car)
library(ggplot2)
library(ggpubr)
```

#Import Data
```{r loading in data}
#loading in data
sample_sites <- read_csv("./data/processed/sample_sites.csv")
taxonomy <- read_csv("./data/processed/taxonomy.csv")
trawl_02_25 <- read_csv("./data/processed/trawl_02_25.csv")
trawl_76_01 <- read_csv("./data/processed/trawl_76_01.csv")
```

#Data Cleaning and Tidying
```{r}
#Separate sample date into day, month year
trawl_76_01 <- trawl_76_01 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_02_25 <- trawl_02_25 %>% 
  separate(SampleDate, into = c('month', 'day', 'year'), sep = "/", remove = FALSE)

trawl_76_01 <- trawl_76_01 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

trawl_02_25 <- trawl_02_25 %>%
  mutate(
    month = factor(
      month,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

combo_trawl <- bind_rows(trawl_76_01, trawl_02_25)
combo_trawl$year <- as.numeric(combo_trawl$year)

```


#Long term Trends of Species

First, I am going to explore long term trends of the data, abundance of species, environmental trends etc. This will help to figure out a direction to go forward with.  

##Total Abundance Shifts over time
Ok first want to look at total abundance has changed over time for all species 
```{r}
annual_abundance <- combo_trawl %>%
  group_by(year) %>%
  summarise(total_fish = sum(Count, na.rm = TRUE))

ggplot(annual_abundance, aes(year, total_fish)) +
  geom_line() +
  theme_minimal()
```
Interestingly a huge spike at the beginning of sampling (sampling effort?) with highs and lows throughout the rest of the years sampled. What would happen if we filter out that first peak?

```{r}
ggplot(subset(annual_abundance, year >1978 ), aes(year, total_fish)) +
  geom_line() +
  theme_minimal()
```
Interesting to see certain peaks more clearly.

##Species richenss changes over time
```{r}
annual_richness <- combo_trawl %>%
  group_by(year) %>%
  summarise(richness = n_distinct(CommonName))

ggplot(annual_richness, aes(year, richness)) +
  geom_line() +
  theme_minimal()
```

##Most abundant species

What are the top 10 most abundant species?
```{r}
top_species <- combo_trawl %>%
  group_by(CommonName) %>%
  summarise(total = sum(Count, na.rm = TRUE)) %>%
  arrange(desc(total)) %>%
  slice(1:10) %>%
  pull(CommonName)

print(top_species)
```


##Unique species over the years
I want to first see how many unique species there are in all of the trawl data. Then I want to look at the abundace of each individual species by year 
```{r}
unique(combo_trawl$CommonName)

#Species
species_by_year <- combo_trawl %>%
  filter(!is.na(CommonName)) %>%
  group_by(year, CommonName) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(year, desc(count))

species_by_year

ggplot(species_by_year, aes(x = year, y = count, color = CommonName)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  theme(
    legend.position = "none")


```
Abundances are so variable by species this is difficult to tell honestly. The magnitude is so different. 


#Long Term Trends -- Environmental Variables 

##Looking at seasonal water temp differences
```{r}
ggplot(combo_trawl, aes(x = month, y = WaterTemp)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  labs(
    x = "Month",
    y = "Water Temperature"
  ) +
  theme_minimal()

seasonal_means <- combo_trawl %>%
  group_by(month) %>%
  summarise(
    avg_WaterTemp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(seasonal_means, aes(x = month, y = avg_WaterTemp, group = 1)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    title = "Average Seasonal Cycle of Water Temperature",
    x = "Month",
    y = "Average Water Temperature"
  ) +
  theme_minimal()


combo_trawl <- combo_trawl %>%
  mutate(
    season = case_when(
      month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      month %in% c("Mar", "Apr", "May") ~ "Spring",
      month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      month %in% c("Sep", "Oct", "Nov") ~ "Fall"
    )
  )

ggplot(combo_trawl, aes(x = season, y = WaterTemp)) +
  #facet_wrap(~year) +
  geom_boxplot(fill = "tan", alpha = 0.7)

```

##Are seasonal water temps sig different?
```{r}
seasonal_water_temp_avg <- combo_trawl %>%
  group_by(year, season) %>%
  summarise(
    avg_WaterTemp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

anova_waterseason1 <- aov(avg_WaterTemp ~ season, data = seasonal_water_temp_avg)
summary(anova_waterseason1)
```

##How has water temp changed over time 

```{r}
annual_water_temp <- combo_trawl %>%
  group_by(year) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE)
  )

ggplot(annual_water_temp, aes(x = year, y = mean_temp)) +
  geom_line(color = "firebrick", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_point(color = "firebrick") +
  theme_minimal()
```
Let's look at how trends have changes over years
```{r}
#Changing the sample date column to recognize as date - ugh
combo_trawl$SampleDate <- as.Date(combo_trawl$SampleDate, format = "%m/%d/%Y")

#Get DOY for mapping
combo_trawl <- combo_trawl %>%
   mutate(DOY = yday(SampleDate))

#avg temp per day since multiple sampling per day
daily_water_temp <- combo_trawl %>%
  group_by(year, DOY) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(daily_water_temp, aes(x = DOY, y = mean_temp, group = year, color = factor(year))) +
  geom_line(alpha = 0.7) +
  theme_minimal()

ggplot(daily_water_temp, aes(DOY, mean_temp, group = year)) +
  geom_line(color = "gray70", alpha = 0.5) +
  geom_line(
    data = subset(daily_water_temp, year >= 2020),
    aes(color = factor(year)),
    size = 1
  ) +
  theme_minimal()

```
Ok interesting but a little crazy, let's try by decade. I want to see if peaks or falls are happening sooner or later than previously.

```{r}

#Get decade for mapping
combo_trawl <- combo_trawl %>%
  mutate(
    decade = floor(year / 10) * 10
  )

decade_water_temp <- combo_trawl %>%
  group_by(decade, DOY) %>%
  summarise(
    mean_temp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(decade_water_temp, aes(DOY, mean_temp, color = factor(decade))) +
  geom_line(size = 0.7) +
  theme_minimal()
```


##Ok 70s and 80s looks like they didn't really sample in the summer months? Is that true?
```{r}
# Count number of samples per decade and month
sampling_coverage <- combo_trawl %>%
  group_by(decade, month) %>%
  summarise(samples = n(), .groups = "drop")

# Plot sampling frequency
ggplot(sampling_coverage, aes(x = month, y = samples, fill = factor(decade))) +
  geom_col(position = "dodge") 
```
Ok so 1990 and on have the most sampling --  maybe we should focus on those?

##Flow Patterns Seasonally
```{r}
#Daily flow patterns (I think)
daily_flow <- combo_trawl %>%
  group_by(year, season, SampleDate) %>%
  summarise(
    daily_flow = mean(FlowmeterDifference, na.rm = TRUE),
    .groups = "drop"
  )

#Seasonal flow averages
seasonal_flow <- daily_flow %>%
  group_by(year, season) %>%
  summarise(
    mean_flow = mean(daily_flow, na.rm = TRUE),
    max_flow  = max(daily_flow, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(seasonal_flow, aes(x = season, y = mean_flow, fill = season)) +
  geom_boxplot() +
  theme_minimal() 
```
##FLow Patterns Annually
```{r}
annual_flow <- seasonal_flow %>%
  group_by(year) %>%
  summarise(total_flow = sum(mean_flow, na.rm = TRUE))

ggplot(annual_flow, aes(x = year, y = total_flow)) +
  geom_line(color = "blue") +
  geom_point() +
  theme_minimal()
```


#Salmonid Focus
Now I want to focus specifically on salmonids

##salmonids over years
```{r}
salmonids_by_year <- combo_trawl %>%
  filter(!is.na(CommonName)) %>%
  filter(CommonName == "steelhead" | CommonName == "Chinook Salmon") %>% 
  group_by(year, CommonName) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(year, desc(count))

salmonids_by_year

#Chinook by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="Chinook Salmon"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

#Steelhead by year
ggplot(salmonids_by_year %>% 
         filter(CommonName =="steelhead"), aes(x = year, y = count)) +
  geom_line() +
  geom_point() +
  theme_minimal()

```
##looking at avg water temp by year
```{r}
#Cal avg water temp by year
avg_temp_by_year <- combo_trawl %>%
  group_by(year) %>%
  summarise(avg_temp = mean(WaterTemp, na.rm = TRUE),
            .groups = "drop")

#Filter for chinook & steelhead individually
ch_by_year <- salmonids_by_year %>% 
  filter(CommonName == "Chinook Salmon")
sh_by_year <- salmonids_by_year %>% 
  filter(CommonName == "steelhead")

#temp/ salmon plot data
temp_ch_data <- left_join(ch_by_year, avg_temp_by_year, by = "year")
temp_sh_data <- left_join(sh_by_year, avg_temp_by_year, by = "year")

#Scale for dual axis plotting chinook
scale_factor_ch <- max(temp_ch_data$count, na.rm = TRUE) / 
                max(temp_ch_data$avg_temp, na.rm = TRUE)
temp_ch_data <- temp_ch_data %>%
  mutate(temp_scaled = avg_temp * scale_factor)

#Scale for dual axis plotting steelhead
scale_factor_sh <- max(temp_sh_data$count, na.rm = TRUE) / 
                max(temp_sh_data$avg_temp, na.rm = TRUE)
temp_sh_data <- temp_sh_data %>%
  mutate(temp_scaled = avg_temp * scale_factor)


#Dual axis plot chinook 
ggplot(temp_ch_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = temp_scaled, color = "Avg Water Temp"), 
            linetype = "dashed", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor,
                        name = "Average Water Temperature")
  ) +
  
  theme_minimal() 

#Dual axis plot steelhead 
ggplot(temp_sh_data, aes(x = year)) +
  geom_line(aes(y = count, color = CommonName), size = 1) +
  geom_line(aes(y = temp_scaled, color = "Avg Water Temp"), 
            linetype = "dashed", size = 1) +
  scale_y_continuous(
    name = "Chinook Count",
    sec.axis = sec_axis(~ . / scale_factor,
                        name = "Average Water Temperature")
  ) +
  
  theme_minimal() 


```





#Looking at avg water temp by year and location
```{r}
avg_temp_table1 <- trawl_76_01 %>%
  group_by(year, Location) %>%
  summarise(
    avg_WaterTemp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

avg_temp_table1


ggplot(avg_temp_table1, aes(x = as.numeric(year), y = avg_WaterTemp, color = Location)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
   geom_smooth(
      aes(group = 1),
      method = "lm",
      se = TRUE,
      color = "black",
      linewidth = 1.2,
      linetype = "dashed"
    ) +
  labs(
    title = "Average Water Temperature Over Time by Location",
    x = "Year",
    y = "Average Water Temperature"
  ) +
  theme_minimal()

avg_temp_table2 <- trawl_02_25 %>%
  group_by(year, Location) %>%
  summarise(
    avg_WaterTemp = mean(WaterTemp, na.rm = TRUE),
    .groups = "drop"
  )

avg_temp_table2

ggplot(avg_temp_table2, aes(x = as.numeric(year), y = avg_WaterTemp, color = Location)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
   geom_smooth(
      aes(group = 1),
      method = "lm",
      se = TRUE,
      color = "black",
      linewidth = 1.2,
      linetype = "dashed"
    ) +
  labs(
    title = "Average Water Temperature Over Time by Location",
    x = "Year",
    y = "Average Water Temperature"
  ) +
  theme_minimal()


```



#Let's start map exploration 
Basic Map of Sample Sites
```{r}
library(sf)
library(terra)
library(dplyr)
library(spData)
install.packages("spDataLarge", repos = "https://nowosad.github.io/drat/", type = "source")

library(spDataLarge)
install.packages("tmap", repos = c("https://r-tmap.r-universe.dev","https://cloud.r-project.org"))
library(tmap)    # for static and interactive maps
library(leaflet) # for interactive maps
library(ggplot2) # tidyverse data visualization package
library(ggthemes)
library(rnaturalearth) #more base maps

library(osmdata) #waterway data
library(elevatr) #elevations
library(nhdplusTools) #NHD datasets

```
##BB of coordinates of delta
```{r}
# Bounding box coordinates (min lon, min lat, max lon, max lat)
delta_bbox <- st_bbox(c(xmin = -121.25, ymin = 37.47, xmax = -122.41, ymax = 39.37),
                      crs = st_crs(4326))

delta_bbox_sf <- st_as_sfc(delta_bbox)
```

##Mapping water
```{r}
#waterways
water_osm <- opq(bbox = delta_bbox) %>%
  add_osm_feature(key = "waterway") %>%
  osmdata_sf()

# extract rivers, canals, etc
rivers <- water_osm$osm_lines
```

```{r}
#lakes 
lakes_osm <- opq(bbox = delta_bbox) %>%
  add_osm_feature(key = "natural", value = "water") %>%
  add_osm_feature(key = "water",
                  value = c("lake", "reservoir", "bay", "lagoon"),
                  match_case = FALSE) %>%
  osmdata_sf()

water_polygons <- lakes_osm$osm_polygons
```

##Map of waterways
```{r}
ggplot() +
  geom_sf(data = st_crop(water_polygons, delta_bbox),
          fill = "lightblue", color = NA) +
  geom_sf(data = st_crop(rivers, delta_bbox),
          color = "blue", size = 0.3) +
  coord_sf(xlim = c(delta_bbox$xmin, delta_bbox$xmax),
           ylim = c(delta_bbox$ymin, delta_bbox$ymax)) +
  theme_minimal()
```

```{r}
sample_sites_sf <- st_as_sf(
  sample_sites,
  coords = c("Longitude", "Latitude"),
  crs = 4326
) 
```

##Map of sample sites
```{r}
ggplot() +
  geom_sf(data = st_crop(water_polygons, delta_bbox),
          fill = "lightblue", color = NA) +
  geom_sf(data = st_crop(rivers, delta_bbox),
          color = "blue", size = 0.3) +
  coord_sf(xlim = c(delta_bbox$xmin, delta_bbox$xmax),
           ylim = c(delta_bbox$ymin, delta_bbox$ymax)) +
  geom_sf(data = sample_sites_sf, color = "red", size = 2) +
  theme_minimal()
```
Ok that was difficult! I would like to find a way to reduce the amount of water shown, focusing primarily on bays estuaries and major rivers and figure out a better bb for the delta.

Moving on...





